---
title: "Historic Maps"
subtitle: "Dataset of the Week (Week 4)"
output: 
  rmdformats::readthedown:
    toc_depth: 3
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}

library(here)
library(tidyverse)

```


# Background

The Harvard Map Collection has a large number of historic, paper
maps. You can visit them in person and a librarian will help you
find a map that interests you and scan a copy of it. Make an 
appointment from this page: 

[Map Collection Appointments](https://outlook.office365.com/owa/calendar/HarvardMapCollection1@HU.onmicrosoft.com/bookings/){target="_blank"}

I encourage you to visit the Map Collection in groups of three or four 
students (with related interests) so that we don't overwhelm 
available appointments with our class.

Historic maps are a source of spatial data that can offer insight into
what a place was like in the past, how it was perceived by those who 
made maps, and what mapmakers wished to communicate about places. 

The maps that you create today serve a specific purpose, and this is 
true of historic maps as well. 

# Historic Maps of London (1854, 1898, 1914)

These three historic maps of London were each created at a 
different time for a different purpose.

## John Snow's Cholera Map (1854)

John Snow was an English doctor the time of an 1854 Cholera 
outbreak in the Broad Street region of London. Based on interviews
with households with Cholera cases and his observations of water from
a public pump located on Broad Street, he believed that the
pump was the source of the outbreak. He created a map indicating 
the number of Cholera deaths at each address, and also showing the
locations of public pumps. This map helped convince local officials
to removed the handle from the Broad Street pump, which was followed 
by an immediate, sharp decline in Cholera cases.

```{r, echo=FALSE, out.width = '100%'}
here::here("week6",
     "old-maps",
     "Snow-cholera-map-1.jpg") |>
  knitr::include_graphics()
```

## Charles Booth's Poverty Maps (1898)

Charles Booth was a successful London businessman in the Victorian era, and he was 
interested in the conditions of the poor. In 1884, while working on a project
related to distributing funds for poverty relief, he became very unsatisfied
with the usefulness of available census data for purposes of understanding
poverty. In 1885, he came across work by socialist activist Henry Hyndman,
which suggested that one in four Londoners lived in extreme poverty. He 
believed this was a gross overstatement of the extent of the problem, and 
the following year, he began work on his own detailed survey, which 
involved interviews with factory owners, school boards, and clergy.

The results was a detailed set of twelve map sheets that color-coded each
street in London into one of seven categories:

* Lowest class. Vicious, semi-criminal
* Very poor, casual. Chronic want.
* Poor. 18s to 21s a week for a moderate family.
* Mixed. Some comfortable, others poor.
* Fairly comfortable. Good ordinary earnings.
* Middle class. Well-to-do.
* Upper-middle and Upper classes. Wealthy.

He ultimately concluded that Hyndman's estimate was likely an understatement.

```{r, echo=FALSE, out.width = '100%'}
here::here("week6",
     "old-maps",
     "booth.png") |>
  knitr::include_graphics()
```

## Max Gill's Wonderground Map of London (1914)

In the early twentieth century, Frank Pick, general manager of the London
Underground commissioned several artists to design
subway posters as part of a publicity campaign. Among these
was a map by MacDonald (Max) Gill that came to be known as the *Wonderground
map of London*. 

The text around the border of the map says, "By paying us your pennies, you
go about your business in trams, electric trains, and motor-driven buses, in 
this largest of all cities, London by the Thames."

```{r, echo=FALSE, out.width = '100%'}
here::here("week6",
     "old-maps",
     "Gill.jpg") |>
  knitr::include_graphics()
```

# W.E.B. Du Bois's Data Portraits (1900)

In 1900, Thomas Calloway and Booker T. Washington persuaded the United
States government to include an exhibit at the *Exposition Universelle* in 
Paris about the progress made by African Americans since Emancipation, 
which was in addition to other exhibits celebrating America's industrial
and political strength. 
Calloway invited W.E.B. Du Bois to contribute a social study of 
African American Life to the exhibit. Du Bois and his students at 
Atlanta University prepared two sets of infographics, which included maps.
One of these focused specifically on Georgia, and the other focused on 
more national and global scales.^[Battle-Baptiste, Whitney, and Britt Rusert, eds. 2018. W. E. B. Du Bois’s Data Portraits: Visualizing Black America. Amherst, Massachusetts New York: Princeton Architectural Press.]

The following three maps represent towns and counties in Georgia.
Each map shows the spatial distribution of Black residents,
represented as dots with colors (red, green, blue, and black) 
to classify people by social class. Yellow dots represent to approximate
distribution of white residents.

## Albany, Dougherty County, Georgia

```{r, echo=FALSE, out.width = '100%'}
here::here("week6",
     "old-maps",
     "du-bois-1.png") |>
  knitr::include_graphics()
```

## McIntosh County, Georgia (outside of Darien)

```{r, echo=FALSE, out.width = '100%'}
here::here("week6",
     "old-maps",
     "du-bois-2.png") |>
  knitr::include_graphics()
```

## Darien, McIntosh County, Georgia 

```{r, echo=FALSE, out.width = '100%'}
here::here("week6",
     "old-maps",
     "du-bois-3.png") |>
  knitr::include_graphics()
```