---
title: "Skills: Digitizing historic map data"
subtitle: "Week 6"
output: 
  rmdformats::readthedown:
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#knitr::knit_engines$set(python = reticulate::eng_python) 
```

This page introduces useful skills for extracting spatial data from 
historic maps

# Creating a georeferenced raster layer from an image

If you have an image file from a scanned map (for example, a historic
map, or one you or a collaborator drew/annotated by hand), you can
import it into GIS software and georeference it, which associates each
pixel in your image with a geospatial location.

In ArcGIS, you'll want to start by loading the image you want to 
georeference. You can add image files from the "Add Data" button.
If the image hasn't been georeferenced, it will probably place it
on your map at about zero degrees latitude, zero degrees longitude 
(in the ocean just west of Africa).

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
library(here)

here("week6",
     "Examples",
     "gifs",
     "arcgis-load-image.gif") |>
  knitr::include_graphics()
```

Zoom your map to the approximate area covered by your map image, then 
click on the "Georeference" button on the "Imagery" tab.

Click "Add Control Points". You'll need to select at least three points on the
image to georeference it. Ideally, there will be spaced far apart
from one another (e.g. near the sides of the image) and will form 
a triangle (rather than being along a straight line).

Select a point on the image, followed by the corresponding location 
on the basemap (you may need to turn off the display on the image to see the basemap). Repeat this at least three times. Each time you add a 
control point the map will shift, rotate, or stretch to minimize the
distances between the points on the image and the points on the map.

When you are satisfied with the position of the image, click "Save"
and then "Close Georeference."

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week6",
     "Examples",
     "gifs",
     "arcgis-georef.gif") |>
  knitr::include_graphics()
```

# Manually creating a layer of data

Once you've loaded a georeferenced layer, you might want to features
to create a vector layer.

In ArcGIS< you'll need to start by creating a new feature class. You'll
use the "Create Feature Class" tool from the Geoprocessing toolbox. 
Specify a name and a geometry type for your new feature class, then 
click "Run." You'll see a new layer added to the contents pane, but
it will have no data in it.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week6",
     "Examples",
     "gifs",
     "arcgis-make-feature class.gif") |>
  knitr::include_graphics()
```

Now you can add features to your new feature class. Click the "Create"
button on the "Edit" ribbon. In the "Create Features" tool, select
the layer you want to add to, and add your features by clicking on the
map. When you are done, click the "Finish" button (the green check mark).

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week6",
     "Examples",
     "gifs",
     "arcgis-make-points.gif") |>
  knitr::include_graphics()
```
