---
title: "Skills: Digitizing historic map data"
subtitle: "Week 6"
output: 
  rmdformats::readthedown:
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#knitr::knit_engines$set(python = reticulate::eng_python) 
```

This page introduces useful skills for extracting spatial data from 
historic maps

# Creating map tiles from a digital map held by a library

AllMaps is an online tool for creating XYZ map tiles from digital 
maps from a libary's online collection. A digital maps held by a 
library will typically have unique url called an International
Image Interoperability Framework (IIIF) Manifest
associated with it. Here's how you find a the IIIF Manifest for a map
from the Harvard Map Collection:

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
library(here)

here("week6",
     "Examples",
     "gifs",
     "manifest-link-harvard.gif") |>
  knitr::include_graphics()
```

And here is how you would find the IIIF manifest for a map from 
the Library of Congress:

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week6",
     "Examples",
     "gifs",
     "manifest-link-loc.gif") |>
  knitr::include_graphics()
```

Once you've copied the link for your map to your clipboard, 
navigate your browser to https://editor.allmaps.org/#/

Paste the Manifest URL into the input box and click the georeference
button. Your browser window will show your image on the left and a
web map on the right. 

Click a point on the image, followed by the corresponding location 
on the web map. You'll need to select at least three points on the
image to georeference it. Ideally, there will be spaced far apart
from one another (e.g. near the sides of the image) and will form 
a triangle (rather than being along a straight line).

Once you are satisfied with your control points, click the 
"Results" button. It will display a url for the map tiles it
has created based on your georeferenced map. Copy this to your
clipboard.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
# here("week6",
#      "Examples",
#      "gifs",
#      "allmaps.gif") |>
#   knitr::include_graphics()
```


# Loading map tiles

If you've created map tiles using AllMaps, you can load this
into a GIS software of your choice.

## ArcGIS

In ArcGIS, you can load XYZ tiles by selecting "Add data from path"
from the "Add Data" menu. Paste the URL you copied from AllMaps into
the text box and click the "Add" button. The layer will appear on your
map.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week6",
     "Examples",
     "gifs",
     "arcgis-load-tiles.gif") |>
  knitr::include_graphics()
```

## QGIS

In QGIS, you can load XYZ tiles by clicking "Add XYZ layer" from the
"Add Layer" item on the Layer menu. This is the same approach 
you took to adding an OpenStreetMap basemap. 

Click the "New" button. Enter the URL from AllMaps and give your
layer a name. Click "OK" and you'll be able to see this new
layer on the dropdown menu and add it to your map.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week6",
     "Examples",
     "gifs",
     "qgis-add-tiles.gif") |>
  knitr::include_graphics()
```

## R

You can load XYZ tiles to an interactive leaflet map as well.

```{r}
library(leaflet)

leaflet() |>
  setView(lng = -77.02357271537899,
                   lat = 38.9033843204262, 
                   zoom = 14) |>
  addTiles("https://allmaps.xyz/maps/7e99290bfc28e845/{z}/{x}/{y}.png")
```

