---
title: "Skills: Network Distances with OpenStreetMap data"
subtitle: "Week 4"
output: 
  rmdformats::readthedown:
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#knitr::knit_engines$set(python = reticulate::eng_python) 
```

This page introduces useful skills for calculating network distances
using network data from OpenStreetMap.

The R examples presented here use the following packages:

```{r, message=FALSE}
library(sf)
library(here)
library(tidyverse)
library(knitr)
library(osmextract)
library(crsuggest)
library(units)
```

It also uses the `r5r` package. Prior to loading the `r5r` package, you
should increase R's memory allocation to Java.

```{r, message=FALSE}
options(java.parameters = '-Xmx2G')

library(r5r)
```

OpenStreetMap is has the ambitions of being a complete map of the world, 
but data in some parts of the world are more complete than in others.
The examples on this page are for Porto Alegre, Rio Grande do Sul, Brazil.

# Getting OpenStreetMap data

You can download OpenStreetMap data directly from R, QGIS, or 
ArcGIS. You can also download the data from an interactive, browser-based
tool.

## Interactive browser-based tools

Two easy-to use extract services are run by BBBike and by the
Humanitarian OpenStreetMap Team (HOT).

### BBBike

Navigate your web browser to https://extract.bbbike.org/. 

Select a format for your data. If you want to work with your data within
ArcGIS, select "Shapefile (Esri). If you want to use it to build a routeable
street network in R (using the `r5r` package), you should select "Protocol buffer (PBF)".

Type in the name of the area you are searching for. This can be anything -- it's 
mostly just for your own reference. If you click on the "search" link, you can zoom to a
a selected area on the map, and fill in the name of that area in the text box.

Enter your email address. You will receive and an email when your data are ready
to download.

Click the "Here" button to display a bounding box on the map. Your data extract
will include all the OpenStreetMap data within this rectangle. If you are satisfied 
with the bounding box, click the "Extract" button to start creating your data extract.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week4",
     "examples",
     "gifs",
     "bbbike.gif") |>
  knitr::include_graphics()
```

### HOT OSM

Navigate your browser to https://export.hotosm.org/en/v3/

You will need to create a free OpenStreetMap account the first time
you use this service.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week4",
     "examples",
     "gifs",
     "osm-account.gif") |>
  knitr::include_graphics()
```

Once you have an account, you can click the "Start
Exporting" button to start creating an extract. You may need to 
log into your account and click a button to allow the 
export tool to access your OSM account.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week4",
     "examples",
     "gifs",
     "hot-login.gif") |>
  knitr::include_graphics()
```

You can use the search bar to define the area for your data extract, 
or you can use one of the tools on the map pane to define the area

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week4",
     "examples",
     "gifs",
     "hot-area.gif") |>
  knitr::include_graphics()
```

The first tab "1 Describe" asks you to describe your extract. This 
is for your own reference (and for anyone else who might want
to use the extract you're making, if you publish it).

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week4",
     "examples",
     "gifs",
     "hot-describe.gif") |>
  knitr::include_graphics()
```

On the next tab, you can specify a data format. You can select 
multiple formats. If you want to create a routable network 
in R, *and* also have a shapefile or geojson file to work
with, you might choose to create a *.pbf file and another 
format.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week4",
     "examples",
     "gifs",
     "hot-formats.gif") |>
  knitr::include_graphics()
```

On the third tab, you can select the features you want to export
(filtering them from the full map based on key/value tags). In this example, 
I'm creating an extract that includes all roads, schools, and hospitals.

Then, on the final tab, you can confirm all the items you've specified
and click the "Export" button. You'll receive an email (to the address
associated with your OSM account) when the data are ready to download.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week4",
     "examples",
     "gifs",
     "hot-filter-export.gif") |>
  knitr::include_graphics()
```

## R

The `osmextract` and `osmdata` packages will both download load OpenStreetMap
data directly into R. The examples here use the `osmextract` package because
it will download data in the `*.pbf` format that the `r5r` package
requires. `osmdata` downloads data as OSM XML files, but it does 
give you a little more flexibility in 
filtering by key/value tags and defining a precise bounding box, so it's 
a good package to be aware of. Keep in mind that converting from
\*.osm files to \*.pbf files can be done with a tool like 
[osmconvert](https://wiki.openstreetmap.org/wiki/Osmconvert).

It will often be helpful to start with a geospatial layer defining the 
boundary of the area you're interested in. I've downloaded the 
municipal boundaries for Brazil from https://data.humdata.org/dataset/cod-ab-bra
and I will load those and filter them to get the Porto Alegre boundaries.

The coordinate reference system is not defined, but the coordinates look
like latitude/longitude coordinates to me, so I'm assuming its WGS 84.

```{r, results='hide', message=FALSE}
poa_boundary <- here("week4",
                 "examples",
                 "bra_admbnda_adm2_ibge_2020") |>
  st_read() |>
  filter(ADM2_PT == "Porto Alegre") |>
  st_set_crs("WGS84")
```

To get all the OSM data for a particular place, use the function `oe_get` from 
the `osmextract` package. It will search for a previously created extract for the
location you've specified. If it can't find it, it sometimes return a larger area.

For example in this case, I've asked for an extract for Porto Alegre and I can see
a message that, since there is not pre-defined extract for that city, it's going
to give me the entire South Region of Brazil (Sul). I can add a `boundary`
parameter to filter the output to only include the area within the Porto 
Alegre boundary (actually a rectangular bounding box containing the Porto 
Alegre boundary). The `oe_get()` function can return points or lines. 

```{r, results='hide'}
poa_osm_points <- oe_get("Porto Alegre",
                         layer = "points",
                         boundary = poa_boundary)
```

Once I have all the points in Porto Alegre, I can filter that data to 
find all the schools and all the hospitals. `osmextract` was written primarily
with transportation applications in mind, so the amenities tags are included
in a column called `other_tags`. You can search within that column
for schools and hospitals.

I've already filtered my extract to include only the area within Port
Alegre's bounding box, but now I'm filtering it further to only include
points within the municipal boundary.

```{r}
poa_schools <- poa_osm_points|>
  filter(str_detect(other_tags, pattern = '"amenity"=>"school"')) |>
  st_filter(poa_boundary)

poa_hospitals <- poa_osm_points|>
  filter(str_detect(other_tags, pattern = '"amenity"=>"hospital"')) |>
  st_filter(poa_boundary)
```

When I load all the streets, I'll also want to save the pbf file
in a directory I can use later to create my routeable street network, 
so I'm speficying a `download_directory`.

I'm going to use `st_intersection` rather than `st_filter` this time to
get only the portions of roads that fall within the municipal boundary.
Note that `st_intersection` is editing the geospatial data that I've loaded 
in R, but isn't editing the *.pbf file I've saved to my `download_directory`.

```{r, results='hide'}

poa_osm_lines <- oe_get("Porto Alegre",
                         layer = "lines",
                         boundary = poa_boundary,
                         download_directory = here("week4",
                                                   "examples")) |>
  st_intersection(poa_boundary)
```

You'll find that most of the line features are roads (features with a highway key), but
if you want, you can filter out the ones that aren't.

```{r}
poa_roads <- poa_osm_lines |>
  filter(!is.na(highway))
```

Now we can take a quick look at the data we've loaded.

```{r}
ggplot() +
  geom_sf(data = poa_boundary) +
  geom_sf(data = poa_roads,
          color = "gray") +
  geom_sf(data = poa_schools,
          color = "steelblue") +
  geom_sf(data = poa_hospitals,
          color = "darkred") +
  theme_void()
```

## QGIS

The QUickOSM plugin will let you load OpenStreetMap data directly to QGIS.
You can access this tool from the Vector menu. If you don't see it there, 
you may need to enable or install the plugin.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week4",
     "examples",
     "gifs",
     "add-quick-osm.gif") |>
  knitr::include_graphics()
```

Once you open the QuickOSM tool, you can specify the key/value tags and the 
place you want to get data from. Click the "Run Query" button and the layers will
be added to your map.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week4",
     "examples",
     "gifs",
     "quick-osm.gif") |>
  knitr::include_graphics()
```

## ArcGIS

Ricardo Klinger has created a very helpful tool for ArcGIS tool called OSMquery that
will import OpenStreetMap data for you. You can find the GitHub repository
for this tool here: https://github.com/riccardoklinger/OSMquery

Download the tool by navigating to its repository and downloading the entire thing
as a zip file.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week4",
     "examples",
     "gifs",
     "download-osoquery.gif") |>
  knitr::include_graphics()
```

Save the zipped file in or near your ArcGIS project directory, and unzip it.

From ArcGIS Pro, open the catalog pane. Right click on "Toolboxes", and select
"Add Toolbox." Navigate to the file you unzipped and select the file called 
"OSMQuery.pyt".

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week4",
     "examples",
     "gifs",
     "add-osmquery.gif") |>
  knitr::include_graphics()
```

The new toolbox you've added includes two scripts: "Get OSM Data" and "Get OSM Data (Expert Tool)".
Select the "Get OSM Data" script. You'll be able so specify one key (e.g. highway or amenity) and
one or more value tags for that key. You can specify the area you want data from by specifying
a bounding box or by entering the name of a place.

After you click "Run" the tool may run for several minutes (I've edited the waiting time
out of the demonstration below) and then it will add layers for any points, lines, and
polygons that match the parameters you've specified.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week4",
     "examples",
     "gifs",
     "run-osmquery.gif") |>
  knitr::include_graphics()
```

# Creating a routable street network

Once you've a layer of roads from OpenStreetMap, you'll need to 
create a routeable network that can be used to calculate paths
between pairs of points.

## R

We'll be calculating paths in R using the `r5r` package. You'll
need to have a folder in your RStudio project that includes
a *.pbf file that includes all the roads in your study area.

Then you'll pass the location of that directory into the `setup_r5()`
function. This will create a couple new files in your directory,
including one called
network.dat, which the `r5r` package will use to calculate network 
distances. 

```{r, message=FALSE, warning=FALSE}
my_core <- here("week4",
                "examples",
                "poa-network") |>
  setup_r5()
```

## QGIS

In QGIS, you can use a plugin called "ORS Tools" to calculate network distances
this will calculate distances directly from the 
[openrouteservice](https://openrouteservice.org/){target="_blank"} API (which
is based on OpenStreetMap data), so there is no need to create your own network within QGIS.

Install the "ORS Tools" plugin from the QGIS plugin manager. Once it's installed, 
you'll find it under the "Web" menu in QGIS.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week4",
     "examples",
     "gifs",
     "web-menu.png") |>
  knitr::include_graphics()
```

You will need to set up a free account and request an API key for openrouteservice.
Navigate your browser to https://openrouteservice.org/dev/#/login and click the 
"Sign Up" button. You'll need to create a username, provide your name and an email
address, and accept the terms of service. 

Once you've created your account and logged in, you'll see a screen with a "Profile"
tab and a "Tokens" tab. Go to the "Tokens" tab and request a token. Once you've generated
the token, you'll see it listed on your "Tokens" tab. Copy it to your clipboard.

In QGIS, select "Provider Settings" from the ORS Tools menu and paste your API
token into the API Key text box.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week4",
     "examples",
     "gifs",
     "ors-api-key.gif") |>
  knitr::include_graphics()
```

## ArcGIS

Creating a network from OpenStreetMap data in ArcGIS Pro gets pretty complicated, 
but ArcGIS Online has its own global, routeable street network that ArcGIS Pro
will use by default. This is generally going to be fine for most purposes (and may
be better in some places than OpenStreetMap data).

# Generating isochrones

An isochrone is a polygon with edges defined by the area you can reach 
within a particular length of time. 

## R

The `r5r` package doesn't generate isochrones directly, but you can create them
by calculating the travel time between a set of points defining locations of interest
and a each cell in a grid that covers your entire study area. Then you can combine
cells that are reachable within your defined threshold.

### Generate a grid

You can use the `st_make_grid()` function to create a grid that covers
the extent of your study area. 

You'll want to create the grid using a projected coordinate system 
by casting the study area boundary to the first suggested CRS from 
the `crsuggest` package, and then immediately transform it to WGS 84.

By default, `st_make_grid` creates a square grid, and you can create
a hex grid by setting `square = FALSE`. I'm setting the size of the 
grid cells to 500 meters. This is probably a coarser grid than what
you would want for a walkshed, but I'm setting it this big so that
you can see the individual grid cells in this demonstration.

I also need to make this layer into an `sf` data frame with a column
called `id` (this is requirement for the `r5r` package).

```{r, message=FALSE}
proj_crs <- suggest_top_crs(poa_boundary)

hex_grid <- st_make_grid(st_transform(poa_boundary, proj_crs),
                         cellsize = set_units(500, m),
                         square = FALSE) |>
  st_transform("WGS84") |>
  st_as_sf()

hex_grid <- hex_grid |>
  mutate(id = seq(1, nrow(hex_grid), by=1))
```

Here's what the resulting grid looks like:

```{r}
ggplot() +
  geom_sf(data = hex_grid) +
  geom_sf(data = poa_boundary, fill = NA, color = "blue") +
  theme_void()
```

### Calculate a travel time matrix

The `r5r` package requires all input layers to have a column 
called `id` to uniquely identify each feature. So I'll add that
to my layer of school points.

```{r}
poa_schools <- poa_schools |>
  mutate(id = seq(1, nrow(poa_schools), by=1))
```

I want to calculate the travel time from each school to the
centroid of each cell in my grid, so I'll create a set of 
grid centroids.

```{r, warning=FALSE}
grid_pts <- hex_grid |>
  st_centroid()
```
Now I can use the `travel_time_matrix` function from `r5r` to 
calculate the distance from every school to every point on my grid. Once you've
finished this step, you should run `stop_r5` to close down the Java
program that's running in the background.

```{r, warning=FALSE, message=FALSE}
travel_times <- travel_time_matrix(my_core,
                                   origins = poa_schools,
                                   destinations = grid_pts,
                                   mode = "WALK")

stop_r5()
```

The result will be a table with three columns:

* `from_id`: The ID for the origin (schools)
* `to_id`: The ID for the destination (a point on the grid)
* `travel_time_p50` the estimated travel time from the origin 
to the destination.

### Filter based on threshold travel times

We can filter out the rows with travel times greater than the thresholds
we want for our isochrones (let's say 20 minutes).

```{r}
travel_times <- travel_times |>
  filter(travel_time_p50 < 21) 
```

Now we can filter our grid to only include the cells with 
IDs that remain in our travel time table.

```{r}
hex_grid_20min <- hex_grid |>
  filter(id %in% as.numeric(travel_times$to_id))
```

Here is the part of the hex grid that is within
20 minutes of a school.

```{r}
ggplot() +
  geom_sf(data = hex_grid_20min) +
  geom_sf(data = poa_schools) +
  theme_void()
```

### Combine grid cells into isochrones

Finally, we can combine those grid cells into a single 
isochrone.

```{r}
iso_schools <- hex_grid_20min |>
  st_union()

ggplot() +
  geom_sf(data = iso_schools) +
  geom_sf(data = poa_schools) +
  theme_void()
```


## QGIS

Open the OSR Tools dialog. From the Batch Jobs tab, click the
"Isochrones from Layer" button. Set the travel mode to 
"foot-walking" and the set the threshold travel time in the 
box that says "Comma-separated ranges."

Click run. The tool might run for several minutes (edited out of the
demonstration below), and may return a lot of "OverQueryLimit" messages,
which you can ignore.

This will generate a separate polygon for each point feature.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week4",
     "examples",
     "gifs",
     "generate-isos-qgis.gif") |>
  knitr::include_graphics()
```

You can use the Dissolve geoprocessing tool to dissolve these into a 
single multipolygon.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week4",
     "examples",
     "gifs",
     "dissolve-isos.gif") |>
  knitr::include_graphics()
```

## ArcGIS

ArcGIS Pro refers to an isochrone as a service area. 

Click the Network Analysis button on the Analysis ribbon and
select "Service Area". You'll notice that there is also an option
to set the Network Data Source to something other than the one
from ArcGIS Online, if you have one (for example, if you've taken
the time to create one from OpenStreetMap data, which we won't be
demonstrating here).

This will create an empty service area layer (you'll see the service
area features in your contents pane, but with nothing displayed on the
map).

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week4",
     "examples",
     "gifs",
     "service-area-1.gif") |>
  knitr::include_graphics()
```

Add the layer of points you want isochrones for as "Facilities" in your 
service area layer. Go to the "Service Area Layer" ribbon and click the 
"Import Facilities" button. Select the point layer you're using from 
the "Input Locations" dropdown menu and click "Apply" or "Okay."

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week4",
     "examples",
     "gifs",
     "import-facilities.gif") |>
  knitr::include_graphics()
```

Now, from the Service Area Layer ribbon, you can also specify the mode of transportation
your isochrones are using, the direction of travel (which shouldn't matter for walking, 
but can matter for motorized modes), the threshold travel time, 
and the time of day (again, shouldn't matter for walking).
You can also specify whether you want the isochrones to dissolve into one another or to 
be a separate polygon for each point. Once you're satisfied with these settings, you can
click "Run".

You will see a set of polygons added to your Service Area layer.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week4",
     "examples",
     "gifs",
     "generate-isos.gif") |>
  knitr::include_graphics()
```

