---
title: "U.S. Census TIGER files"
subtitle: "Dataset of the Week (Week 1)"
output: 
  rmdformats::readthedown:
    toc_depth: 3
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}

library(here)
library(tidyverse)

```


# Background

The United States Census Bureau maintains and publishes geographic 
data on the United States and its territories. These data are published
as TIGER files. TIGER stands for **T**opologically **I**ntegrated 
**G**eographic **E**ncoding and **R**eferencing.

You can read the Census Bureau's description of its mission, authority,
and activities here:
[https://www.census.gov/about/what.html]{target="_blank"}.

They describe their mission as "to serve as the nation's leading
provider of quality data about its people and economy."

The Census Bureau uses
TIGER files to create maps and to collect and summarize data about the 
people and economy of the United States.

# Available data

Three types of TIGER files are available: _geographic areas_
and _features_. 

Geographic areas will be polygons. Features may be points, 
lines, or polygons. Two useful variables that will be 
included with any TIGER polygon layer are `ALAND` and
`WATER`, which represent the land and water areas of the 
polygon respectively, in square meters (this can be useful
when you want to calculate a population density based on the
land area rather than on the total area of a region). 

## Geographic areas

Geographic areas are areas for which the Census Bureau
publishes demographic data. They may be nested or non-nested.

### Nested geographies

Many national statistical agencies, including the United States
Census Bureau, use nested geographies for tabulating population
and demographic data. The nested geographies the United States
Census Bureau uses include:

* States (and state equivalents)
* Counties (and county equivalents)
* Census tracts
* Census block groups
* Census blocks

We describe these as nested because the smaller geographies
nest within larger geographies. Census block groups comprise
a set of census blocks; census tracts comprise a set of census
block groups; counties comprise a set of census tracts; and
so on.

Census geographic areas can be identified by a numeric identifier
called a GEOID (geographic identifier) or FIPS (Federal Information
Processing System) code. These codes are nested as well. The FIPS code
for a *census block* is a 15 digit-code. The first 12 digits of the
code are FIPS the FIPS code for the *block group* the block is in. 
The first 11 digits of the block group FIPS code are the FIPS code
for the *tract*. Figure 1 illustrates the structure of 
FIPS code for nested census geographies.

```{r, echo=FALSE, fig.cap="Figure 1: FIPS code for nested census geographies", out.width = '50%'}

here("images",
     "FIPS.png") |>
  knitr::include_graphics()
```

**States (and state equivalents)** include 50 states, five territories, 
and the District of Columbia. State boundaries are political 
boundaries and they were drawn for political and historical reasons.
State boundaries are quite stable over time. 

Each state is divided into one or more **counties (or county equivalents)**.
The District of Columbia and Guam each have one county equivalent. Texas
has 254 counties. Like state boundaries, county boundaries were 
generally drawn
for political and historical reasons. They may change occasionally, and 
these changes are more common than changes to state boundaries.^[In 
June 2022, the boundaries for county equivalents in Connecticut changed
substantially. See https://portal.ct.gov/Office-of-the-Governor/News/Press-Releases/2022/06-2022/Governor-Lamont-Announces-Census-Bureau-Approves-Proposal-for-Planning-Regions]

Counties are divided into **census tracts**. Census tracts are not 
political boundaries. The Census Bureau established census tract
boundaries based on established 
criteria, including (among other considerations) a goal of 
maintaining a similar population size in each tract.^[For detailed criteria 
used to determine tract boundaries for the 2020 census, see 
https://www.federalregister.gov/documents/2018/11/13/2018-24567/census-tracts-for-the-2020-census-final-criteria] Tract boundaries can change as often
as every ten years.

Census tracts are divided into **census block groups**. Block group
boundaries are likewise drawn by the Census Bureau based on 
established criteria^[See https://www.federalregister.gov/documents/2018/11/13/2018-24570/block-groups-for-the-2020-census-final-criteria] and may change as often as 
every ten years.

Census block groups are sets of adjacent **census blocks**. Census
blocks are bounded by physical features such as roads (most commonly), 
rivers, or coastlines. 

### Non-nested geographies

Other census geographies are not nested, including:

* Places (including cities)
* Urbanized areas
* Core-based statistical areas
* Public-use microdata areas

FIPS codes for non-nested census geographies are assigned 
alphabetically.

Most cities and towns in the United states are classified
as **places**. This a category that includes both 
legally-incorporated places (e.g. cities and towns) and unincorporated 
places that are equivalent to incorporated places and have a locally-used
name, but do not have an functioning government.^[See https://www.census.gov/programs-surveys/bas/information/cdp.html] Many 
locations in the United States
are not classified as being part of any incorporated or unincorporated 
place. About 24 percent of land area of Massachusetts belongs to
incorporated or unincorporated places. In California, less than ten 
percent of all land is part of an incorporated or unincorporated place.

The places TIGER file inclused a variable `LSAD` that differentiates 
among types of places, including the following categories:

* City (LSAD = 25)
* Township (LSAD = 35)
* Town (LSAD = 43)
* Village (LSAD = 47)
* Census-designated place (unincorporated) (LSAD = 57)

An **Urban area** is group of contiguous census blocks that 
have a population density above an established threshold, irrespective
of municipal or county boundaries.^[See https://www.federalregister.gov/documents/2022/03/24/2022-06180/urban-area-criteria-for-the-2020-census-final-criteria for details.] 

**Core-based statistical areas (CBSAs)** are groups of *counties* that share
economic ties, which are primarily determined based on identified 
commuting patterns.^[Given recent changes in the prevalence of 
commuting, this may change. See https://www.federalregister.gov/documents/2021/07/16/2021-15159/2020-standards-for-delineating-core-based-statistical-areas for a description of
current criteria, including a discussion of comments on them.] Many counties 
in the United States are not part of any core-based statistical area.

A core-based statistical area may be either a 
**metropolitan statistical area (MSA)** or a 
**micropolitan statistical area ($\mu$SA)**. An MSA is a group
of connected counties containing at least one urban area with a 
population of at least 50,000 people. A $\mu$SA us a group
of connected counties that are not in an MSA but contain
at least one urban area with a population of at least 10,000 
people. The CBSA TIGER file differentiates between MSAs and 
$\mu$SA with a variable called `LSAD`, which takes a value of `M1`
for MSAs and `M2` for $\mu$SAs. There are 392 MSAs in the United
States and 547 $\mu$SAs.

A **public-use microdata area (PUMA)** is a set of contiguous census
tracts within a single state that have a combined population of
at least 100,000 people. Each census tract in the United States
is classified into one (and only one) PUMA. To protect privacy, 
individual- and household-level data from the United States census 
are not available with any geographic identifiers for areas smaller 
than a PUMA.

## Features

Features may be represented on a census map for context, and may
define the boundaries of geographic areas. They include:

* Roads
* Rails
* Water
* Coastline
* Landmarks

Features will have an MTFCC (MAP/TIGER Feature Class Code) to 
classify each feature in the layer as one of several types. This
document includes definitions of all current MTFCCs: https://www2.census.gov/geo/pdfs/reference/mtfccs2022.pdf
    
### Roads 

Roads represent the roadway centerline with MTFCC codes to 
differentiate among categories including:

* Primary Road (S1100)
* Secondary Road (S1200)
* Local neighborhood Road, Rural Road, City Street (S1400)
* Walkway/Pedestrian Trail (S1710)

A file with only primary roads is available as a single 
national file. Files with only primary and secondary 
roads are available at the state level. Files including
all road features are available for individual counties.

### Rails

All rail features have the same MTFCC. Rail data is available as a line
layer in a single national file. 

### Water

Water features are available as linear features or as
polygon features. 

**Linear water features** have MTFCCs to classify them into categories including:

* Stream/river (H3010)
* Canal, Ditch, or Aqueduct (H3020)
* Connector (H1100)

**Polygon water features** have MTFCCs to classify them into categories including:

* Lake/pond (H2030)
* Stream/river (H3010)
* Bay/Estuary/Gulf/Sound (H2051)

Water feature layers are available for each county.

### Coastline

The coastline is available as a line layer in a single national
file with only one MTFCC value.

### Landmarks

Landmarks are available as point data or as polygon data for each state. 

For point landmarks, MTFCCs differentiate among categories including:

* Mountain peak or summit (C3022)
* Cemetery (K2582)
* School or academy (K2543)
* Government center (K2165)
* Library (K2195)
* Golf course (K2561)
* Fire department (K2193)
* Place of worship (K3544)

For polygon landmarks, MTFCCs differentiate among categories including:

* Island (C3023)
* Hospital/Hospice/Urgent Care Facility (K1231)
* Park (K2180)
* Cemetery (K2582)
* University or College (K2540)
* Local Jail or Detention Center (K1236)

# Data access

Two ways to access TIGER files are through an interactive, web-based 
interface and through an the census bureau's API.

## Web interface

The easiest way to find TIGER files in a web browser is to enter
"United States Census TIGER files" into a search engine. The top result
is likely to be https://www.census.gov/geographies/mapping-files/time-series/geo/tiger-line-file.html

Scroll down to see a set of tabs indicating different years. You may
find a note that "The 2023 TIGER/Line Shapefiles will be available in September 2023."
Select the tab for 2022 (or any earlier year) and click on the "Web Interface" link.

This will take you to https://www.census.gov/cgi-bin/geo/shapefiles/index.php (or you could
just navigate directly to that site). From there, instructions to find and download
the shapefile you want are fairly self-explanatory. A "submit" button will always
lead to a screen with more options; a "download" button will download the data.

## API access

An application programming interface (API) allows software developers 
to access data programmatically (i.e. through code). The R package
`tigris` offers a convenient way for R users to access the Census 
Bureau's TIGER API directly from their code. The Python package
`pygris` (from the same developer) is similar.

You can find full set of `tigris` functions to download data 
here: https://walker-data.com/census-r/census-geographic-data-and-applications-in-r.html#data-availability-in-tigris

Here are some examples of how to use the `tigris` package to
load the datasets described above into R.

```{r, message=FALSE, results='hide', warning=FALSE, eval=FALSE}
library(tigris)

# load all state boundaries in the United States
states <- states()

# Load all county boundaries in the United States
counties <- counties()

# Load all county boundaries in Massachusetts
TX_counties <- counties(state = "Massachusetts")

# Load all census tract boundaries in Rhode Island
RI_tracts <- tracts(state = "Rhode Island")

# Load all census tract boundaries in Los Angeles County
LA_tracts <- tracts(state = "California", county = "Los Angeles")

# Load all census block groups in Rhode Island
Suffolk_BGs <- block_groups(state = "RI")

# Load all census block groups in Cook County, Illinois
Cook_Co_BGs <- block_groups(state = "IL", county = "Cook")

# Load all census blocks in Rhode Island and Connecticut
RI_blocks <- blocks(state = "RI")

# Load all places in Montana
places <- places(state = "Montana")

# Load all urban areas in the United States
UAs <- urban_areas()

# Load all CBSAs in the United States
CBSAs <- core_based_statistical_areas()

# Load all PUMAs in New York state
PUMAs <- pumas(state = "New York")

# Load all primary roads in the United States
US_roads <- primary_roads()

# Load all primary and secondary roads in New Hampshire
NH_pri_sec_roads <- primary_secondary_roads(state = "NH")

# Load all roads in Fulton County, Georgia
fulton_roads <- roads(state = "Georgia", county = "Fulton")

# Load all rail lines in the United States
rails <- rails()

# Load all linear water features in Salt Lake County, Utah
slc_water_line <- linear_water(state = "UT", county = "Salt Lake")

# Load all polygon water features in King County, Washington
king_co_water_poly <- area_water(state = "WA", county = "King")

# Load the coastlines for the United States
coast <- coastline()

# Load the point landmarks for Massachusetts
MA_landmarks_pt <- landmarks(state = "MA")

# Load the polygon landmarks for Guam
guam_landmarks_poly <- landmarks(state = "Guam", type = "area")
```
