---
title: "Skills: Varying feature symbology"
subtitle: "Week 2"
output: 
  rmdformats::readthedown:
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#knitr::knit_engines$set(python = reticulate::eng_python) 
```

This page introduces useful skills for displaying
geospatial data on a map using R, Python, QGIS, or ARcGIS.

The R skills on this page use the following R packages:

```{r, message=FALSE}
library(sf)
library(here)
library(tidyverse)
library(tidycensus)
library(ggspatial)
library(ggthemes)
library(RColorBrewer)
library(knitr)
library(tigris)
```


# Finding census variables

There are so many census variables to choose from, finding the exact ones
you need is not at all trivial.

## R

The easiest way to download census data if you're working in R is to
use the `tidycensus` package.

The `load_variables` function will allow you to search for the variables
that interest you. It takes two required parameters: The year, and the
dataset. 

For the Decennial Census years before 2020, use `"sf1"`. 

The full data from the 2020 Decennial Census haven't been released 
yet, but limited data are available in the redistricing data. To see these
variables, use `"pl"` as your dataset.

For the American Community Survey, use `"acs1"` for one-year estimates
or `"acs5"` for five-year estimates.

If you run the line below in your console, you'll see this table describing 
all the available variables from the 2010 Decennial Census. Note that 
this is searchable.

```{r, eval=FALSE}
load_variables(2010, "sf1") |> View()
```

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
library(here)

here("images",
     "load_vars.png") |>
  knitr::include_graphics()
```

## Web interface

Here's how you would find 2010 Decennial Census variables for 
total population by race for census tracts in Wayne County, Michigan.

Navigate your browser to https://data.census.gov
and select "Advanced Search."

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("images",
     "census-search.png") |>
  knitr::include_graphics()
```

Set the geography filter to all census tracts in Wayne County.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week2",
     "examples",
     "gifs",
     "web-tracts.gif") |>
  knitr::include_graphics()
```

Set the survey filter to the Decennial Census (Summary File 1) and
the year filter to 2010.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week2",
     "examples",
     "gifs",
     "web-tvy-yr.gif") |>
  knitr::include_graphics()
```

Under the Topic filter, select "Race and Ethnicity", and check the 
"Race and Ethnicity" box. Then click the "Search" button on the bottom
right corner of the browser.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week2",
     "examples",
     "gifs",
     "topic-filter.gif") |>
  knitr::include_graphics()
```

The top result in this case is Table P9: "Hispanic or Latino, and not Hispanic
or Latino, by Race."

# Downloading census data

Once you've identified the variables you want, you can download them.

## R

Once you've identified the names of the variables you want, you can use `get_acs` to get data from the 
American Community Survey or `get_decennial` to get data from the Decennial census.

For either of these functions, you need to specify the year and the geography for which you 
want data. For some small geographies, you'll need to specify a state and possibly a county.
You'll also specify the list of variable names you need.

By default, the result will be a table with the name of the variable in one column
and the value in another, like this:

```{r, message=FALSE}
CDs_HI_long <- get_decennial(geography = "congressional district", 
                        state = "Hawaii", 
                        year = 2010,
                        variables = c("H004001", 
                                      "H004004"))

CDs_HI_long
```

You can specify that you want each variable in its own column by setting
`output = 'wide'`.

```{r, message=FALSE}
CDs_HI_wide <- get_decennial(geography = "congressional district", 
                        state = "Hawaii", 
                        year = 2010,
                        variables = c("H004001", 
                                      "H004004"),
                        output = 'wide')

CDs_HI_wide
```

You can also indicate that you want the data as a geospatial dataset that includes the boundaries
the the geographic areas by setting `geometry = TRUE`.

```{r, message=FALSE, results='hide'}
counties_DE_geo <- get_decennial(geography = "county", 
                        state = "Delaware", 
                        year = 2010,
                        variables = c("H004001", 
                                      "H004004"),
                        output = 'wide',
                        geometry = TRUE)
```

```{r}
ggplot(counties_DE_geo) +
  geom_sf(aes(fill = H004001)) +
  scale_fill_viridis_c() +
  theme_void()
```

Here is how I would get the race and ethnicity data for census tracts in 
Wayne County, Michigan.

```{r, warning=FALSE, message=FALSE}
race_vars <- c(total_pop = "P009001",
               hisp_lat = "P009002",
               nh_white_alone = "P009005",
               nh_black_alone = "P009006",
               nh_am_in_alone = "P009007",
               nh_asian_alone = "P009008",
               nh_pac_is_alone = "P009009",
               nh_other_alone = "P009010",
               nh_multiracial = "P009011")

wayne_race <- get_decennial(geography = "tract",
                            variables = race_vars,
                            year = 2010,
                            state = "MI",
                            county = "Wayne",
                            output = "wide")
```


## Web interface

Continuing the previous example, where we identified a table with race and
ethnicity data for all census tracts in Wayne County, Michigan: 

When I select this table, I see that it's organized with census tracts
in columns and variables in rows. It will be more useful to 
have variables in columns and census tracts in rows, and you can 
switch to this format by selecting the "Transpose" button.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week2",
     "examples",
     "gifs",
     "transpose.gif") |>
  knitr::include_graphics()
```

Then you can download the data as a csv file.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week2",
     "examples",
     "gifs",
     "save.gif") |>
  knitr::include_graphics()
```

# Cleaning/reshaping a dataset

If you download the data using R or Python, it will be organized in a way 
that makes it easy to bring into a GIS program. This is not necessarily the
case for data you download through https://data.census.gov/.

## Microsoft Excel

If you downloaded your dataset as a csv from https://data.census.gov/,
it will look something like this:

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("images",
     "census-excel.png") |>
  knitr::include_graphics()
```

Data values are in every other row, and some of the data cells include 
parenthetical notes.

### Remove rows without values

You can sort the data to separate rows with blank values in a column, then
delete the blank rows.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week2",
     "examples",
     "gifs",
     "delete-blank-rows.gif") |>
  knitr::include_graphics()
```

### Remove notes within cells

You can remove the text notes in some cells by using the 
"Text-to-columns" wizard available from the "Data" tab.
This will take everything after a space and remove it from the
column.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week2",
     "examples",
     "gifs",
     "delete-notes.gif") |>
  knitr::include_graphics()
```

### Extract unique part of tract name

To join this data to the census tract boundaries TIGER file, you'll want to 
extract the unique part of the tract name. You can use a similar process to 
the one you used to delete the notes. 

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week2",
     "examples",
     "gifs",
     "split-tract-name.gif") |>
  knitr::include_graphics()
```

### Remove white space

There are also some spaces at the beginning of each tract name, and these
will cause you problems when you want to match these values with the tract
names in a TIGER file. To make matters worse, these are not normal spaces, 
so some simple methods (like the `TRIM()` function in Excel) won't work here.
So here's what you do.

First, insert a couple of columns.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week2",
     "examples",
     "gifs",
     "insert-cols.gif") |>
  knitr::include_graphics()
```

Next, identify the specific kind of space character this is using the 
`CODE()` function in Excel.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week2",
     "examples",
     "gifs",
     "id-white-space.gif") |>
  knitr::include_graphics()
```

Now you can use the `SUBSTITUTE()` function in Excel to replace this 
weird white space with nothing (empty quotes).

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week2",
     "examples",
     "gifs",
     "sub-white-space.gif") |>
  knitr::include_graphics()
```

Finally, paste these values into their own column (as values). Once you have those,
you can delete the original column and the column with the `SUBSTITUTE` formulas.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week2",
     "examples",
     "gifs",
     "paste_vals.gif") |>
  knitr::include_graphics()
```

# Joining non-spatial data to spatial data

You may sometimes find you have a table of non-spatial data that 
references places by an identifier or place name, and a separate 
spatial dataset that indicates that locations of those places. 

## R

```{r, echo=FALSE, message=FALSE, results='hide'}
wayne_tracts <- tigris::tracts(state = "MI", county = "Wayne", year = 2010)
```

Here is an example where I have a dataframe called `wayne_race` that 
includes populations in several race/ethicity categories for each 
census tract in Wayne County, Michigan, but it does not include 
a geospatial column with the tract boundaries.

I have a separate dataframe called `wayne_tracts` (which I downloaded
using the `tigris` package) that includes the census tract boundaries.

I can join these into a single dataframe based on the value of a common 
variables (a variable that is present in both dataframes). Both dataframes
have a variable with each tract's FIPS code. In `wayne_race`, the FIPS code is in
a variable called `GEOID`. In `wayne_tracts`, it's in a variable called `GEOID10`.
I'll start by renaming one of these variables to match the other.

Notice that the `rename()` function uses the form `new_name = old_name`. This
tends to trip my up a lot, since I would have found the opposite to be 
more intuitive.

```{r}
wayne_tracts <- wayne_tracts |>
  rename(GEOID = GEOID10)
```

Now I can use `left_join()` to join these two dataframes.

```{r, message=FALSE}
wayne_race_tracts <- wayne_tracts |>
  left_join(wayne_race)
```

`wayne_race` has 11 variables. `wayne_tracts` has 15 variables. `wayne_race_tracts` has
25 variables: All 15 from `wayne_tracts` plus the 10 new variables from `wayne_race`.

## QGIS

In QGIS, add the non-spatial data table by clicking the "Open Data Source Manager" button. While you can add a table by dragging a csv file directly only your
map window, this will tend to add numeric data as text, which will cause problems.
This method will correctly read text as text.

From the data source manager, select "Delimited text". Specify the file name
of the file, select the radio button for csv files, and the one to indicate 
that this is a data table with no geometry. In the sample data, confirm that
QGIS has selected the correct data type (integers for count data), then select
"Add" and "Close".

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week2",
     "examples",
     "gifs",
     "add-csv-qgis.gif") |>
  knitr::include_graphics()
```

Open both attribute tables to
see the names of the variables that shares the same values between the two 
tables.

Double-click the geospatial layer to and select the button for "Manage joins 
to other layers" along the side of the Layer Properties dialog (it looks like 
a little circle next to a bigger triangle).

Click the plus sign at the bottom of the dialog, then use the drop-down 
menus to indicate what table you're joining data from, and the variable names
of the common variable. Then click Okay.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week2",
     "examples",
     "gifs",
     "join-qgis.gif") |>
  knitr::include_graphics()
```

## ArcGIS

To join data from a table to a geospatial layer, you'll need to start by
adding both the shapefile with the geospatial data and the data file (in this
case, a csv file) to you map.

You can add a csv file to your map the same way you would add a shapefile. It
won't display anything on your map, but it will appear in you Contents 
pane under "Standalone Tables".

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week2",
     "examples",
     "gifs",
     "arcgis-add-table.gif") |>
  knitr::include_graphics()
```

Open the attribute table for the geospatial layer
as well as the non-spatial data table. Find the name of the column 
in each table that will be used to match rows.

Then, open the "Add join" geoprocessing tool. You'll indicate the 
geospatial layer as the input table and the data table as the 
join table. For each, you'll indicate which column to use for the
join. Then click "Run."

After it runs, the attribute table for the geospatial layer will include
all the data from the data table.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week2",
     "examples",
     "gifs",
     "join-arcgis.gif") |>
  knitr::include_graphics()
```

# Calculating a new variable value

You might want to combine two or more variables to calculate
a new variable.

## R

The `mutate()` function in R is for calculating or setting values
for new variables.

In this example, I have a dataframe with the total population and the
land area (in square meters) of each census tract in Wayne County, Michigan.

Note: The `head` function displays the first few rows of a dataset
and the `kable()` function displays the table in a neat format. These
are useful for me putting together a tutorial, but aren't necessary for 
any kind of analysis.

```{r}
wayne_pop_area <- wayne_race_tracts |>
  select(total_pop, ALAND10)

head(wayne_pop_area) |>
  kable()
```

I can use the `mutate` function to calculate the population per 
square kilometer. I need to divide the population by the land area, and
also multiply by one million to convert from square meters to square
kilometers.

```{r}
wayne_pop_area <- wayne_pop_area |>
  mutate(density = 1000000 * total_pop / ALAND10)

head(wayne_pop_area) |>
  kable()
```


## QGIS

Open the attribute table for a layer and click the edit button. Then click the
"Open Field Calculator" button. 

In the Field Calculator dialog, enter a name for the new variable, and indicate
the variable type (e.g. integer or decimal). Enter an expression for calculating
the new variable. To find and add existing variable names, expand the list
under "Fields and Values". Once you've entered the formula, click OK. 

To save your changes, click the editing button again to end your edit session.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week2",
     "examples",
     "gifs",
     "field-calc-qgis.gif") |>
  knitr::include_graphics()
```

## ArcGIS

With the attribute table open, you can click the "Add field" button and enter 
a name for the new field.**Note: Make sure you select the right data type. "Long" means the values will be whole numbers, "Double" means they can be decimal values.**

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week2",
     "examples",
     "gifs",
     "add-field-arcgis.gif") |>
  knitr::include_graphics()
```

Right click on the heading of the field you just created and select "Calculate Field." Then you can enter the formula you want to use to calculate the 
values for the new field. As much as possible, I recommend clicking on 
items to add them to your formula rather than typing it in directly.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week2",
     "examples",
     "gifs",
     "calc-field-arcgis.gif") |>
  knitr::include_graphics()
```

# Clipping an area from a polygon

Geographic boundaries like census tracts or counties
will often include area of water, and your analysis may be more concerned
with the land area.

## R

Here are the census tracts in Wayne County, Michigan and the water areas
within the county. Some tract boundaries extend beyond the coastlines of
Lake Erie and Lake St Clair. 

```{r, message=FALSE, results='hide'}
wayne_water <- area_water(state = "MI",
                          county = "Wayne") |>
  st_union()
```

```{r, message=FALSE, warning=FALSE}
ggplot(wayne_pop_area) +
  geom_sf(data = wayne_water,
          color = NA,
          fill = "lightblue") +
  geom_sf(color = "gray25",
          fill = NA) +
  theme_void()
```

The st_difference function can clip all the areas from a polygon layer that are
within another polygon layer

```{r, warning=FALSE}
clipped_tracts <- wayne_pop_area |>
  st_difference(wayne_water)

ggplot(clipped_tracts) +
  geom_sf() +
  theme_void()
```

Specific to United States census geography, the `tigris` package has a handy
function called 

```{r, message=FALSE, warning=FALSE}
no_water_tracts <- wayne_pop_area |>
  erase_water()

ggplot(no_water_tracts) +
  geom_sf() +
  theme_void()
```

## QGIS

In QGIS, the Clip tool will keep the parts of a layer that are within
another layer. The Difference tool will keep the parts of a layer 
that are outside another layer.

From the processing menu, select Toolbox, and search for the Difference 
tool. In this example, I want to exclude water areas from my census tracts, 
so I'm setting the census tracts as my input layer and the water as my overlay
layer. I'm also specifying a filename for my output layer. Then I click Run
and Close and the new layer is added to my map.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week2",
     "examples",
     "gifs",
     "clip-qgis.gif") |>
  knitr::include_graphics()
```

## ArcGIS

Open both the layer you want to clip and the layer you want to use to 
clip it. Select all features in the clipping layer.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week2",
     "examples",
     "gifs",
     "select-all-arcgis.gif") |>
  knitr::include_graphics()
```

From the Edit tab, click the "Modify" button, and then select "Clip" from the
"Modify Features" pane. Check the box for "Clip all editable features" and select "Discard" to clip away the parts of the polygons that overlap the clipping layer.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week2",
     "examples",
     "gifs",
     "clip-arcgis.gif") |>
  knitr::include_graphics()
```

# Creating a chloropleth map

A chloropleth map is in which a region is divided into
subarea, and the color of each subarea represents the
value of a variable.

The techniques for varying the color of polygons by the value of a variable
is covered in the Week 1 skills page, [here](https://c-voulgaris.github.io/VIS-2128/week1/skills.html#setting-symbology-by-attribute).

## More on colors in R (static/ggplot)

A little bit more on colors in R:

Let's start with a chloropleth map of population densities in Wayne County, Michigan.

```{r}
density_map <- ggplot(no_water_tracts) +
  geom_sf(aes(fill = density)) +
  theme_void()

density_map
```

By default, the color ramp for continuous variables ranges from dark blue to light blue.

There are several predefined color palettes you can use, and some useful functions you can use to select colors in `ggplot`.

The [viridis](https://bids.github.io/colormap/){target="_blank"} color ramps are meant to be colorblind friendly and to reproduce reasonably well in grayscale (for example, if someone prints your graphic then photocopies it with a black-and-white printer). There are 4 options.

Option A:

```{r}
density_map +
  scale_fill_viridis_c(option = "A")
```

Option B:

```{r}
density_map +
  scale_fill_viridis_c(option = "B")
```

Option C:

```{r}
density_map +
  scale_fill_viridis_c(option = "C")
```

Option D (this is the default if you don't specify an option):

```{r}
density_map +
  scale_fill_viridis_c(option = "D")
```

Option E:

```{r}
density_map +
  scale_fill_viridis_c(option = "E")
```

Option F:

```{r}
density_map +
  scale_fill_viridis_c(option = "F")
```

Option G:

```{r}
density_map +
  scale_fill_viridis_c(option = "G")
```

Option H:

```{r}
density_map +
  scale_fill_viridis_c(option = "H")
```

You can reverse the color ramp values by setting `begin = 1, end = 0`.

```{r}
density_map +
  scale_fill_viridis_c(option = "H", 
                       begin = 1, end = 0)
```

You can also set up your own color ramp using `scale_fill_gradient()` and specifying low and high values. Here's as scale that varies continuously from blue to red.

```{r}
density_map +
  scale_fill_gradient(low = "blue", high = "red")
```

`scale_fill_gradient2()` will let you specify low, medium, and high values. Here's a ramp that varies from green, to yellow, to red (which would be intuitive because of the analogy to a traffic signal, but not accessible for someone with red-green colorblindness). You also need to set a midpoint value - which I'll set as the median value.

```{r}
density_map +
  scale_fill_gradient2(low = "green", mid = "yellow", high = "red",
                       midpoint = median(no_water_tracts$density, 
                                         na.rm = TRUE))
```

`scale_fill_gradientn()` will let you specify as many colors as you want to include in your color ramp. Here's a ramp that varies from red to orange to yellow to green to blue to purple.

```{r}
density_map +
  scale_fill_gradientn(colors = c("red", 
                                  "orange", 
                                  "yellow", 
                                  "green", 
                                  "blue", 
                                  "purple"))
```

The `RColorBrewer` package offers a bunch of pre-defined palettes that can be useful for continuous data. You can view the options with the function `display.brewer.all()

```{r, fig.height=8}
library(RColorBrewer)

display.brewer.all()
```

The top group of palettes you see above are sequential palettes, which range from dark to light colors.

```{r}
map_palette <- brewer.pal(5, "PuBuGn")

density_map +
  scale_fill_gradientn(colors = map_palette)
```

The last group of colors are diverging palettes, which range from dark to light to dark colors.

```{r}
map_palette <- brewer.pal(5, "Spectral")

density_map +
  scale_fill_gradientn(colors = map_palette)
```

The middle group are qualitative palettes which work well for categorical variables, but would be confusing for a continuous color ramp.

There are some other fun package that define color palettes, including [wesanderson](https://github.com/karthik/wesanderson){target="_blank"} (colors inspired by Wes Anderson films), [PNWColors](https://jakelawlor.github.io/portfolio/pnwcolors/){target="_blank"} (colors inspired by the package author's photos of the Pacific Northwest), [MetBrewer](https://github.com/BlakeRMills/MetBrewer/tree/main){target="_blank"} (inspired by works at the Metropolitan Museum of Art), and [MexBrewer](https://paezha.github.io/MexBrewer/){target="_blank"} (inspired by works of Mexican muralists). These four packages are structured similarly. Some of the available palettes don't work well for continuous variables (the order of the colors isn't intuitive), but some work well.

```{r}
library(MetBrewer)

degas_contin <- met.brewer(name = "Degas", n = 5, type = "continuous")

density_map +
  scale_fill_gradientn(colors = degas_contin)
```

# Log-transforming a color ramp

Sometimes a variable will have a lot of very low values and
a few very high values. A log transformation on the color ramp
can make a the variation among low values more legible.

## R

```{r, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
la_pop <- get_decennial(geography = "tract",
                            variables = "P009001",
                            year = 2010,
                            state = "CA",
                            county = "Los Angeles",
                            output = "wide")

la_tracts <- tracts(state = "CA",
                    county = "Los Angeles",
                    year = 2010) |>
  erase_water() |>
  rename(GEOID = GEOID10) |>
  left_join(la_pop) |>
  mutate(density = 1000000 * P009001 / ALAND10)
```

In R, you can log-transform the color ramp by setting the parameter
`trans = "log"` withing the `scale_fill_*` function.

Here is a map without a log-transformed color ramp.

```{r}
ggplot(la_tracts) +
  geom_sf(aes(fill = density),
          color = NA) +
  scale_fill_viridis_c(option = "H") +
  theme_void()

```

And here it is with a log-transformed color ramp.

```{r, warning=FALSE}
ggplot(la_tracts) +
  geom_sf(aes(fill = density),
          color = NA) +
  scale_fill_viridis_c(option = "H", 
                       trans = "log") +
  theme_void()
```

## ArcGIS

In ArcGIS, you'd need to create a new version of the variable with 
log-transformed values.

Before doing the calculation, you'll need to filter out any areas with a 
value of zero for the value you want to log-transform (because you'll get 
an error if you try to take the logarithm of zero).

Start by creating a new variable. **Note: Again, make sure you select the right data type. "Long" means the values will be whole numbers, "Double" means they can be decimal values.**

Calculate the value of the new variable as `math.log(!density!)` (if density is
the name of your density variable).

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week2",
     "examples",
     "arc-gis-log-calc.png") |>
  knitr::include_graphics()
```

Set the primary symbology to as "unclassed colors" to get a 
continuous color ramp. If you choose the density variable in the "Field" menu,
you will see default values for the upper and lower labels as the (rounded)
minimum and maximum values for that variable.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week2",
     "examples",
     "arc-gis-dens-ramp-1.png") |>
  knitr::include_graphics()
```

If you change the feild to your log-transformed variable, you can manually
change the labels for minimum and maximum values to correspond to the 
minimum and maximum values for the untransformed density value, rather than
showing the log-transformed values on your legend.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week2",
     "examples",
     "arc-gis-dens-ramp-2.png") |>
  knitr::include_graphics()
```

# Sampling points in a polygon

When you are representing counts within areas on a map, one way to 
visualize them is to randomly distribute points within each area, where each points represents
a specific number of the thing you're counting.

To do this, you may want to generate a number of points at random locations
within a polygon.

## R

In R, you can use the function `st_sample` to sample points within a 
polygon. The first argument of this function is a set of polygons, and 
second argument indicates the number of points you want to generate within
each polygon.

Let's say we want to represent the number of non-Hispanic white and non-Hispanic
Black residents on a map as points on a map, with each point representing
100 residents. Here's how we would create the points.

```{r, warning=FALSE, message=FALSE}
pts_nh_white <- st_sample(wayne_race_tracts,
                          size = round(wayne_race_tracts$nh_white_alone/100))

pts_nh_black <- st_sample(wayne_race_tracts,
                          size = round(wayne_race_tracts$nh_black_alone/100))  
```

And now we can add those points to a map.

```{r}
ggplot(no_water_tracts) +
  geom_sf() +
  geom_sf(data = pts_nh_black,
          color = "darkorange",
          size = 0.1) +
  geom_sf(data = pts_nh_white,
          color = "steelblue",
          size = 0.1) +
  theme_void()
```


## QGIS

In QGIS, you can generate random points within a polygon using the 
"Random Points in Polygon" tool.

To generate one point per (for example) hundred people, you would 
define the number of points per polygon as an expression where you 
divide the value of the population variable by 100.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week2",
     "examples",
     "gifs",
     "random-pts.gif") |>
  knitr::include_graphics()
```

# Dot density ArcGIS shortcut

Rather than creating a separate layer of points, you can create
a dot density map in ArcGIS directly from the symbology pane. 

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week2",
     "examples",
     "gifs",
     "dot-dens-argis.gif") |>
  knitr::include_graphics()
```
