---
title: "Open Street Map (Philippines)"
subtitle: "Dataset of the Week (Week 4)"
output: 
  rmdformats::readthedown:
    toc_depth: 3
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}

library(here)
library(tidyverse)

```


# Background

The goal of OpenStreetMap
is to be a comprehensive map of the entire world. What is worth including
on the map is determined based on the interests and beliefs of OpenStreetMap
contributors, who tend to be well-educated and tech-savvy. Studies have estimated that over ninety percent of OpenStreetMap
contributors are men, although the effect of this gender imbalance on 
the content of contributions is not well-understood (see studies by Das et al.^[Das, Maitraye, Brent Hecht, and Darren Gergle. 2019. “The Gendered Geography of Contributions to OpenStreetMap: Complexities in Self-Focus Bias.” In Proceedings of the 2019 CHI Conference on Human Factors in Computing Systems, 1–14. Glasgow Scotland Uk: ACM. https://doi.org/10.1145/3290605.3300793] and Stephens^[Stephens, Monica. 2013. “Gender and the GeoWeb: Divisions in the Production of User-Generated Cartographic Information.” GeoJournal 78 (6): 981–96. https://doi.org/10.1007/s10708-013-9492-z.]).

Steve Coast founded OpenStreetmap in 2004 when he became frustrated with 
the lack of publicly-available, digitized maps in the United Kingdon. 
He put a GPS receiver in his backpack and started cycling
around with it to create GPS traces representing roads, and used those
traces to create map data. He started presenting his work at conferences
and invited others to contribute to the project of creating an open-source
map of the world.

# Precursors

OpenStreetMap would have 
been inconceivable without the success of Wikipedia and impossible without the
widespread availability of GPS.

Wikipedia had been founded
in 2001, so the idea that loosely-organized volunteer contributors could 
produce a reliable and trustworthy source of information was just beginning 
to take off in 2004. 

The Global Positioning System (GPS) is a project of the United States Department of 
Defense that relies on a constellation of at least 24 satellites that are in orbit around
the earth. Based on information transmitted by these satellites, a GPS reciever
with a clear line of site to at least four satellites can determine its own
precise location on the surface of the earth. The United States started the GPS 
project in 1973 and enabled
civilian access to GPS in the 1980s. The full set of 24 satellites became
operational in 1993. Around that time, the United States initiates a program
called *Selective Availability* that degraded the quality of GPS signals for 
most users (and denied access to specific users). Selective availability ended on
May 20, 2000. From that point on, commercially-available GPS receivers (which 
could be purchased for \$100 - \$200) allowed any person to determine their 
latitude and longitude coordinates with reasonable precision. This was a very 
exciting moment in time.

# Data contributors

The overall structure of the OpenStreetMap project is quite similar to that
of Wikipedia: Anyone can contribute, and anyone can edit another person's 
contributions. 

Local chapters of the OpenStreetMap organization host 
[mapping parties](https://wiki.openstreetmap.org/wiki/Mapping_parties) to create a social environment in which people can (learn how to)
contribute to the map. Volunteers can also propose [mapping projects](https://wiki.openstreetmap.org/wiki/Mapping_projects) organized
around a specific geography or theme and invite others to contribute.

A few institutions have imported large batches of data to OpenStreetMap.

In 2007, a Dutch Navigational software firm (Automotive Navigational Data, which
has since been renamed as GeoJunxion) imported a complete layer of roads 
for the Netherlands.

In 2007 and 2008, all TIGER files from the United States Census Bureau
were imported into OpenStreetMap, which substantially improved the
coverage (if not the accuracy) of OpenStreetMap (the data have required
a lot of cleaning and editing).

In 2010, Microsoft licensed the use of Bing aerial imagery to OpenStreetMaps, 
making it legal for contributors to contribute map data by tracing aerial
imagery (previously, they would have needed to collect GPS traces in person).

Although these contributions by large organizations have been influential,
OpenStreetMap still gives priority to individual mapping volunteers. 

The following video introduces some individuals and organizations who 
have been involved in contributing OpenStreetMap data for the Philippines.

<iframe width="560" height="315" src="https://www.youtube.com/embed/DNGMKJkzEkk?si=JDvPjXixGP-fHvtl" data-external= "1" ></iframe>

# Data format and content

OpenStreetMap data have their own data format, which identifies features
as 

* nodes (points), 
* ways (connected points to form lines or polygons), and 
* relations (related features, like multipolygons or bus stops and associated routes), 

any of which can be identified with an 
unlimited number of key/value tags. These are non-rectangular
datasets, but you can think of a key as a variable name and a value
as the value for that variable. In fact, that's how key/value
tags are treated when OSM data is converted to a rectangular
data format.

Roads (and road-related points like intersections and signs) will always 
be tagged with a [`highway`](https://wiki.openstreetmap.org/wiki/Key:highway) 
key, which can have one of many associated values, including:

* `motorway` (a divided highway)
* `trunk` 
* `primary`
* `secondary`
* `tertiary`
* `unclassified` (a local road that is not residential)
* `residential`
* `footway` (a pedestrian path)
* `bridleway` (a path for horses)
* `steps`
* `corridor` (a hallway within a building)

Other keys that a way with a highway key tag might have include:

* `sidewalk` with possible values:
    * `both` (sidewalks on both sides of the street)
    * `right` (sidewalk only on the right side of the street)
    * `left` (sidewalk only on the left side of the street)
    * `no` (No sidewalk on either side of the street)
    
* `cycleway` with possible values:
    * `lane` (two-way street with bike lanes on both sides)
    * `opposite_lane` (one-way street with a contra-flow lane)

Another common key in OpenStreetMap data is `amenity`, with can have
many associated values, including:

* `bar`
* `cafe`
* `restaurant`
* `kindergarten`
* `school`
* `library`
* `car_wash`
* `parking`
* `baby_hatch` (a place where a baby can be anonymously left)
* `brothel`
* `public_bookcase` (like a Little Free Library)
* `post_box`
* `bench`
* `dog_toilet`
* `shower` (a public shower)
* `kneipp_water_cure` (outdoor foot bath facility)
* `refugee_site` (human settlement for refugees or internally displaced persons)

# Data access

There are a couple of R packages (`osmdata` and `osmextract`) that 
will load OpenStreetMap data in R. QGIS also has a plugin that will
download and load OpenStreetMap data. 

Two sites that make it easy to extract OpenStreetMap data for a 
speficied area are run by:

* BBBike (https://extract.bbbike.org/) and
* Humanitarian OpenStreetMap Team (HOT): (https://export.hotosm.org/en/v3/).

The BBBike extract tool will create a file with all OpenStreetMap
data within a specified bounding box. The HOT-OSM extract tool will
also let you filter for specific key/value tags (useful if, for
example, you're only interested in transportation-related features).
The HOT-OSM extract tool requires you to create a free account with 
OpenStreetMap (which is the same account you would use if you want
to edit OpenStreetMap).




