---
title: "Cape Town Open Data Portal"
subtitle: "Dataset of the Week (Week 3)"
output: 
  rmdformats::readthedown:
    toc_depth: 3
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}

library(here)
library(tidyverse)

```


# Background

The [Cape Town Open Data Portal](https://odp-cctegis.opendata.arcgis.com/){target="_blank"} 
was established in 2016 with the creation of the City's Open Data Policy, 
which you can read here: 
https://resource.capetown.gov.za/documentcentre/Documents/Bylaws%20and%20policies/Open_Data_Policy.pdf

The Open Data Policy lists the following desired outcomes of the open data
portal:

1. Supporting the constitutional right of everyone to access all information 
held by the state.^[The inclusion of this right in the post-Apartheid constitution
may be seen as a reaction to the secretive practices of the Apartheid government.]
2. Increasing public engagement by making it easier for members of the public to
access data.
3. Increasing economic development by supporting the development of technologies
and applications that use public data.
4. Supporting the creation of knowledge through academic research.
5. Enabling integrated planning among city departments.
6. Reducing the administrative burden associated with manually addressing requests
for data.

The purpose of the Open Data Portal is publish data sets that the city has created
for its own purposes and allow others to reuse those data for as-yet unimagined 
purposes.

Cape Town is not unique in its establishment of an Open Data Portal, nor in its
objectives. Many local governments throughout the world have established Open Data
Portals. Cape Town's is good example of a typical effort.

# Relationship with national and international efforts

South Africa is a member of the [Open Government Partnership](https://www.opengovpartnership.org/){target="_blank"}
and has published an [open government action plan](https://www.opengovpartnership.org/wp-content/uploads/2021/01/South-Africa_Action-Plan_2020-2022.pdf){target="_blank"} 
that emphasizes its commitment to a policy of open data. 

# Process for collecting datasets

Cape Town's Information and Knowledge Management (IKM) Department manages
the Open Data Portal. Each city department has a designated data steward 
responsible for cleaning and documenting datasets maintained in their department
and publishing them to the Open Data Portal in cooperation with the IKM
department.

# Supporting software

The Open Data Portal is hosted on the ESRI ArcGIS Online platform.

This is a widely used platform for Open Data Portals. Alternatives include
[DataPress](https://datapress.com/) and [OpenDataSoft](https://www.opendatasoft.com/en/). 

# Accessing the data

The format of the Open Data Portal is similar to that of other
ESRI-based platforms. 

For the most part, geospatial data files are available in any 
of the following data formats:

* csv (without geospatial information for lines or polygons)
* kml
* shapefile
* geojson
* geodatabase