---
title: "Field data and positionality"
subtitle: "Dataset of the Week (Week 3)"
output: 
  rmdformats::readthedown:
    toc_depth: 3
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}

library(here)
library(tidyverse)

```

# Background

This week, we'll be focusing on data that you've collected yourself.
As with all the data sources we've been discussing, data you collect
will reflect the biases and goals of the person or institution (you, in
this case) collecting, curating, and presenting the data. It 
is worth taking a moment to reflect on how your own identity will 
influence the data you will be collecting.

# Your identity

Dimensions of your identity that might influence what you notice and choose
to record about a place might include:

* Race
* Class
* Educational attainment
* Income
* (Dis)ability
* Gender
* Citizenship
* Nationality
* Religion
* Sexual orientation
* Age
* Family relationships and caregiver status

How might you describe your own identity along these or other dimensions?
How might these identities influence your approach to data collection 
for this week's assignment? 

Some possible questions with answers
that might vary depending on your own identity include:

* What signals to you that a place or situation is safe or unsafe?
* What features of a place are noteworthy and what features are so 
normal that they might escape your notice?
* What data-gathering behaviors might draw unwanted attention?
* What sort of attention would be unwanted?
* What does a normal place look like?
* What does an ideal place look like?
* At what point are two things similar enough to be classified as the same thing?

# Your subject community

Any of the above dimensions of your identity might to contribute to your
*insider/outsider* status, which might be a separate, but related dimension: 
Do you feel at home in the place where you are gathering data? Do you consider this 
community to be your own community? Would members of this community
see you as an insider or as an outsider?

How do you view your own position in relation to the community you are studying?

* Are you you collecting data about a community you see yourself as
a part of?
* In what ways do you have lower or higher status than the members of the
community you are studying?
* In what ways do members of the community you are studying have 
more or less power than yourself?
* In what ways do members of the community you are studying have more 
or less relevant expertise than you do?

# Your audience

You will generally collect data that you expect to share with others.
Your beliefs about what your audience will be interested in and what
they will need to know might be based in part of your perceptions
of *their* identities, and your own power, status, and expertise relative
to your own.

* Are you you collecting data for people you see as being like you, 
or for people you see as being different from you?
* Do you view your audience as having higher or lower status than 
yourself?
* Do you view your audience as having more or less power than yourself?
* Do you view your audience as having more or less relevant expertise
than you do?


