---
title: "Skills: Collecting and importing GPX data"
subtitle: "Week 3"
output: 
  rmdformats::readthedown:
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#knitr::knit_engines$set(python = reticulate::eng_python) 
```


This page demonstrates methods for collecting GPS data in the field and importing
it to a GIS software of your choice.

The R methods on this page use the following packages:


```{r, message=FALSE, warning=FALSE}
library(here)
library(sf)
library(tidyverse)
library(ggspatial)
library(lwgeom)
library(maptiles)
library(tidyterra)
library(crsuggest)
```

# Collecting GPS traces

There are many different smart phone applications for both Andorid
and iOS devices that will allow you to collect GPS traces and export
them as GPX files. For this assignment, I recommend the following applications
that make it easy to collect gpx traces and annotated waypoints:

* GPX Tracker ([iOS](https://apps.apple.com/us/app/open-gpx-tracker/id984503772))
* GPS Logger ([Android](https://play.google.com/store/apps/details?id=eu.basicairdata.graziano.gpslogger&hl=en_US))


## GPX Tracker

When you open the app, you'll see a screen that looks like this:

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week3",
     "screenshots",
     "ios-start.PNG") |>
  knitr::include_graphics()
```

When you're ready to start tracking, tap the big green "Start Tracking" button 
at the bottom of the screen. To add a waypoint along the way, tap the button
with a red pin to the left of that big green button. You'll see the pin appear
on your map. By default, the annotation for the waypoint will be the current
time (when you dropped the pin). YOu can edit this by tapping the pin on your map,
and tapping the blue pencil to the left of the current annotation.

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week3",
     "screenshots",
     "ios-edit-name.PNG") |>
  knitr::include_graphics()
```

Then you can replace the default annotation with something else.

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week3",
     "screenshots",
     "ios-edit-name2.PNG") |>
  knitr::include_graphics()
```

When you are done with your trace, tap the save button. Then, tap the reset button
and select the option "Reset" (rather that "Save and Start new").

You can see your saved GPX file by clicking the file icon on the top left of your 
screen. Select the file you have saved and click "Share" to email it to yourself.

## GPS Logger

Go to the Track tab and select tap the start button at the bottom of the screen 
to start recording.

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week3",
     "screenshots",
     "gps-logger-track.PNG") |>
  knitr::include_graphics()
```

Use the Annotate button to add a Waypoint - you will be prompted to enter a
description.

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week3",
     "screenshots",
     "gps-logger-annotate.PNG") |>
  knitr::include_graphics()
```

Tap the stop button when you are finished. The file will appear in your 
tracklist and you will be able to email it to
yourself from there.

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week3",
     "screenshots",
     "gps-logger-tracklist.PNG") |>
  knitr::include_graphics()
```

# Importing GPS traces

Once you've exported a GPX file from a GPS device (such as a smart phone),
you may want to bring it into your GIS software.

GPX files can include data in five layers:

* route_points
* routes
* track_points
* tracks
* waypoints

If you've just collected a GPS trace using GPX Tracker or GPS Logger,
the app was recording your position and the current time at regular intervals (e.g.
every second, or every five seconds). Those times and locations are stored in the
track_points layer. They are connected into a line (based on the order in which they
were recorded) that is stored in the tracks layer. Any annotated waypoints you've
added will be in the waypoints layer.

## R

You can use the `st_read()` function to load data from a GPX file into
R. You will need to specify which layer you want. 

```{r, message=FALSE, results='hide'}
path_line <- here("week3",
              "examples",
              "20250904-154314 - Mutungo biina church route UAZ 565H.gpx") |>
  st_read(layer = "tracks")

waypoints <- here("week3",
              "examples",
              "20250904-154314 - Mutungo biina church route UAZ 565H.gpx") |>
  st_read(layer = "waypoints")
```

And here are track and waypoints displayed on a map.

```{r}
ggplot() +
  geom_sf(data = path_line) +
  geom_sf(data = waypoints,
          color = "red") +
  theme_void()
```

## QGIS

You can import a GPX file to QGIS by dragging and dropping it onto your map.
You'll be able to specify which layers you want. The most relevant layers are
likely to be "waypoints" and "tracks".

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "qgis-import.gif") |>
  knitr::include_graphics()
```

## ArcGIS

You can use the GPX to Features tool (in the Geoprocessing toolbox)
in import a GPX trace to ArcGIS Pro. 

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "arc-import-track.gif") |>
  knitr::include_graphics()
```

When you import the points, you will get the way points and the track points
(the pings that are used to construct the track) in a single layer. 

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "arc-import-pts.gif") |>
  knitr::include_graphics()
```

To have a 
layer of just the waypoints, you'll need to select by attribute (Type is equal to 
WPT) and create
a new layer from those selected features. 

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "arc-select-pts.gif") |>
  knitr::include_graphics()
```

# Split line at specified points

You can use waypoints to split a line into segments.

## R

The `lwgeom` package has several extensions to the `sf` package, including
the `st_split` function. It requires your data to be in a projected coordinate
system, so you'll need to start by transforming your data.

You can use the `suggest_crs` function in the `crsuggest` package to find an 
appropriate projected coordinate 
system for your data.

```{r}

suggest_crs(path_line)

```

From the above list, I'll use UTM zone 36N. The code is 32636. Now I can use
the `st_transform` function to transform my data from latitude/longitude to meters

```{r}
path_line <- path_line |>
  st_transform(32636)

waypoints <- waypoints |>
  st_transform(32636)

```

The st_split function can split a line into components on either side of (one)
point. Here is how I would get the track on either side of a the third of my waypoints.

```{r}

line_parts <- st_split(path_line, waypoints[3,])

```


More to come...

## QGIS

Coming soon...

## ArcGIS

Coming soon...