---
title: "Skills: Collecting and importing GPX data"
subtitle: "Week 3"
output: 
  rmdformats::readthedown:
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#knitr::knit_engines$set(python = reticulate::eng_python) 
```


This page demonstrates methods for collecting GPS data in the field and importing
it to a GIS software of your choice.

The R methods on this page use the following packages:


```{r, message=FALSE, warning=FALSE}
library(here)
library(sf)
library(tidyverse)
library(ggspatial)
library(lwgeom)
library(crsuggest)
library(knitr)
library(units)
```

# Collecting GPS traces

There are many different smart phone applications for both Andorid
and iOS devices that will allow you to collect GPS traces and export
them as GPX files. For this assignment, I recommend the following applications
that make it easy to collect gpx traces and annotated waypoints:

* GPX Tracker ([iOS](https://apps.apple.com/us/app/open-gpx-tracker/id984503772))
* GPS Logger ([Android](https://play.google.com/store/apps/details?id=eu.basicairdata.graziano.gpslogger&hl=en_US))


## GPX Tracker

When you open the app, you'll see a screen that looks like this:

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week3",
     "screenshots",
     "ios-start.PNG") |>
  knitr::include_graphics()
```

When you're ready to start tracking, tap the big green "Start Tracking" button 
at the bottom of the screen. To add a waypoint along the way, tap the button
with a red pin to the left of that big green button. You'll see the pin appear
on your map. By default, the annotation for the waypoint will be the current
time (when you dropped the pin). YOu can edit this by tapping the pin on your map,
and tapping the blue pencil to the left of the current annotation.

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week3",
     "screenshots",
     "ios-edit-name.PNG") |>
  knitr::include_graphics()
```

Then you can replace the default annotation with something else.

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week3",
     "screenshots",
     "ios-edit-name2.PNG") |>
  knitr::include_graphics()
```

When you are done with your trace, tap the save button. Then, tap the reset button
and select the option "Reset" (rather that "Save and Start new").

You can see your saved GPX file by clicking the file icon on the top left of your 
screen. Select the file you have saved and click "Share" to email it to yourself.

## GPS Logger

Go to the Track tab and select tap the start button at the bottom of the screen 
to start recording.

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week3",
     "screenshots",
     "gps-logger-track.PNG") |>
  knitr::include_graphics()
```

Use the Annotate button to add a Waypoint - you will be prompted to enter a
description.

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week3",
     "screenshots",
     "gps-logger-annotate.PNG") |>
  knitr::include_graphics()
```

Tap the stop button when you are finished. The file will appear in your 
tracklist and you will be able to email it to
yourself from there.

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week3",
     "screenshots",
     "gps-logger-tracklist.PNG") |>
  knitr::include_graphics()
```

# Importing GPS traces

Once you've exported a GPX file from a GPS device (such as a smart phone),
you may want to bring it into your GIS software.

GPX files can include data in five layers:

* route_points
* routes
* track_points
* tracks
* waypoints

If you've just collected a GPS trace using GPX Tracker or GPS Logger,
the app was recording your position and the current time at regular intervals (e.g.
every second, or every five seconds). Those times and locations are stored in the
track_points layer. They are connected into a line (based on the order in which they
were recorded) that is stored in the tracks layer. Any annotated waypoints you've
added will be in the waypoints layer.

## R

You can use the `st_read()` function to load data from a GPX file into
R. You will need to specify which layer you want. 

```{r, message=FALSE, results='hide'}
path_line <- here("week3",
              "examples",
              "20250904-154314 - Mutungo biina church route UAZ 565H.gpx") |>
  st_read(layer = "tracks")

waypoints <- here("week3",
              "examples",
              "20250904-154314 - Mutungo biina church route UAZ 565H.gpx") |>
  st_read(layer = "waypoints")
```

```{r, echo=FALSE, results='hide', warning=FALSE}
waypoints <- waypoints |>
  mutate(name = as.numeric(name)) |>
  filter(!is.na(name)) 
```


And here are track and waypoints displayed on a map.

```{r}
ggplot() +
  geom_sf(data = path_line) +
  geom_sf(data = waypoints,
          color = "red") +
  theme_void()
```

## QGIS

You can import a GPX file to QGIS by dragging and dropping it onto your map.
You'll be able to specify which layers you want. The most relevant layers are
likely to be "waypoints" and "tracks".

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "qgis-import.gif") |>
  knitr::include_graphics()
```

## ArcGIS

You can use the GPX to Features tool (in the Geoprocessing toolbox)
in import a GPX trace to ArcGIS Pro. 

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "arc-import-track.gif") |>
  knitr::include_graphics()
```

When you import the points, you will get the way points and the track points
(the pings that are used to construct the track) in a single layer. 

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "arc-import-pts.gif") |>
  knitr::include_graphics()
```

To have a 
layer of just the waypoints, you'll need to select by attribute (Type is equal to 
WPT) and create
a new layer from those selected features. 

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "arc-select-pts.gif") |>
  knitr::include_graphics()
```

# Split line at specified points

You can use waypoints to split a line into segments.

## R

The `lwgeom` package has several extensions to the `sf` package, including
the `st_split` function. It requires your data to be in a projected coordinate
system, so you'll need to start by transforming your data.

You can use the `suggest_crs` function in the `crsuggest` package to find an 
appropriate projected coordinate 
system for your data.

```{r}

suggest_crs(path_line)

```

From the above list, I'll use UTM zone 36N. The code is 32636. Now I can use
the `st_transform` function to transform my data from latitude/longitude to meters

```{r}
path_line <- path_line |>
  st_transform(32636)

waypoints <- waypoints |>
  st_transform(32636)

```

The `st_split` function can split a line into components on either side of (one)
point. It returns a single feature with a geometry collection as its geometry. 
You need to use `st_collection_extract` to get each segment into its own 
row of your dataset.

Here is how I would get the track on either side of a the second waypoint 
(the first was when the surveyor initially boarded the bus, so I wouldn't 
need to split the route there).

I am using the `mutate()` function to create a new variable (called `which`) 
that gives a unique name to the two segments.

```{r}

line_parts <- st_split(path_line, waypoints[2,])|>
  st_collection_extract("LINESTRING") |>
  mutate(which = c("A", "B"))

```

Now I can plot those two segments, differentiating them by color.

```{r}
ggplot() +
  geom_sf(data = line_parts,
          aes(color = which)) +
  theme_void()
```

I can see that the first segment is the one that got labeled as B and the rest
of the route is A (I know that this route was from west to east). 
Now I would want to split the A section into the parts before 
and after the third point, and continue on through all the waypoints. This could
be tedious, but I can write a little loop that will repeat the same set of 
actions over and over until I've worked my way across the line.

First, I'll start a layer called segments that includes the segment I've
already successfully split off, and I'll add to that as I work my way down the 
line. I'll make a new layer that's includes the rest of the line that I want to 
split up.

```{r}
segments <- line_parts |>
  filter(which == "B")

rest_of_line <- line_parts |>
  filter(which == "A")
```

Now here's a loop that works its way through the rest of the waypoints.

```{r}

# repeat the same thing for waypoints 3 through 12
for (i in 3:12) {
  
  # split the line into parts before and after this waypoint
  line_parts <- st_split(rest_of_line, waypoints[i,])|>
    st_collection_extract("LINESTRING") |>
    mutate(which = c("A", "B"))
  
  # Check to see which segment (A or B) touches the last segment we split off.
  next_segment_check <- st_intersects(line_parts, 
                                      segments[nrow(segments),],
                                      sparse = FALSE)
 
  # the side that touches the last segment is the next segment
  next_segment <- line_parts[next_segment_check,]
   
  # add it to the layer of segments
  segments <- rbind(segments, next_segment)
  
  # the other side is the rest of the line
  rest_of_line <- line_parts[!next_segment_check,]
}

```


What's remaining as the rest of the line will be my last segment. And now I can 
give all the segments unique labels (numbering them one through twelve).

```{r}

segments <- rbind(segments, rest_of_line) |>
  mutate(which = 1:12)

```

And now I can see all these separate segments

```{r}

ggplot() +
  geom_sf(data = segments,
          aes(color = as.character(which))) +
  theme_void()

```

## QGIS

To split a line by the points along it, you can install the Split Lines By
Points plugin.

```{r, echo=FALSE, message=FALSE}
here("week3",
     "examples",
     "gifs",
     "qgis-split-plugin.gif") |>
  knitr::include_graphics()
```

You can access the plugin by opening the Toolbox from the Processing menu and
selecting it from there. This will open a dialog where you can specify the line
layer you want to split and teh point layer you want to split it by. You can
(and should) also specify a file you want to save the resulting layer of
segments to. Once you run the tool, it will show up in your table of contents.

It will look like the original line layer, but if you open the attribute table, 
you can see that it's a set of distinct features.

```{r, echo=FALSE, message=FALSE}
here("week3",
     "examples",
     "gifs",
     "qgis-split-run.gif") |>
  knitr::include_graphics()
```

## ArcGIS

In ArcGIS, you can use the Split Line at Point tool to split a line layer based on a layer of points.

```{r, echo=FALSE, message=FALSE}
here("week3",
     "examples",
     "gifs",
     "arcgis-split.gif") |>
  knitr::include_graphics()
```

# Create line from points

As an alternative to splitting the track line into segments, you could also
create a set of lines from the trackpoints that were collected between the times
of your waypoints.

## R

I'll start by loading the track points, and keeping only the variables I need:
time and speed

```{r}

track_points <- here("week3",
                     "examples",
                     "20250904-154314 - Mutungo biina church route UAZ 565H.gpx") |>
  st_read(layer = "track_points") |>
  select(time, speed)

```

I'm going to filter those points to only include the times when the vehicle 
was moving.

```{r}

track_points <- track_points |>
  filter(speed > 0)

```


Here's a quick plot of the track points and waypoints. For the track points, I'm setting the alpha value to 0.1 to 
make them quite transparent.

```{r}

ggplot(track_points) +
  geom_sf(alpha = 0.1) +
  geom_sf(data = waypoints,
          color = "red",
          size = 3) +
  theme_void()

```

My waypoints layer has 20 variables, but the only ones I really need are `time`
and `name` (where my annotations are saved). So I will use the `select()`
function to remove all the columns except just those two (I don't need to 
tell it to keep the geometry column that stores the lat-lon coordinates - it
always will).

```{r}

waypoints <- waypoints |>
  select(time, name)

```

So here's the table of my waypoints (the `kable()` function just makes it 
display pretty).

```{r}

waypoints |>
  kable()

```

To get the set of track points between the first two waypoints, I would want all 
the track points with an earlier time than the second waypoint.

```{r}
seg_1_pts <- track_points |>
  filter(time < waypoints$time[2])
```


Let's see what that looks like.

```{r}

ggplot(seg_1_pts) + 
  geom_sf(alpha = 0.1) +
  geom_sf(data = waypoints,
          color = "red",
          size = 3) +
  theme_void()

```

Then we'll join them into a line (it's a little tricksy, where we cobble together
a couple functions). 

```{r}

seg_1_line <- st_coordinates(seg_1_pts) |>
  st_linestring() |>
  st_sfc(crs = st_crs(track_points)) |>
  st_as_sf() 

ggplot(seg_1_line) + 
  geom_sf() +
  geom_sf(data = waypoints,
          color = "red",
          size = 3) +
  theme_void()

```


## QGIS

TBD

## ArcGIS

TBD

# Calculate line lenth

You might want to be able to say something about the length of your track
(or its respective segments).

## R

In R, you can use the st_length function to calculate length.

```{r}

segments <- segments |>
  mutate(length = st_length(segments)) |>
  mutate(length = set_units(length, "km")) |>
  select(which, length)

kable(segments)

```


## QGIS

In QGIS, you can use the Add geometry attributes tool to calculate length.
You should set your project to a projected coordinate reference system first, 
and be sure you know what units the CRS is in.

```{r, echo=FALSE, message=FALSE}
here("week3",
     "examples",
     "gifs",
     "qgis-calc-length.gif") |>
  knitr::include_graphics()
```

## ArcGIS



# Calculate a cumulative sum

There might be attributes of your waypoints that you want to apply to associated
line segments.

In this example, my first waypoint was when the surveyor first boarded the bus, 
and it's annotated with the number of people who were on the bus when they boarded. 
The subsequent waypoints are annotated with the number of people getting
on (a positive number) or off (a negative number). I can use this to calculate
the total number of passengers on board for the segment following the point
by taking the cumulative sum of the column.

## R

My waypoints layer has 20 variables, but the only ones I really need are `time`
and `name` (where my annotations are saved). So I will use the `select()`
function to remove all the columns except just those two (I don't need to 
tell it to keep the geometry column that stores the lat-lon coordinates - it
always will).

```{r}

waypoints <- waypoints |>
  select(time, name)

```

So here's the table of my waypoints (the `kable()` function just makes it 
display pretty).

```{r}

waypoints |>
  kable()

```

Now I can create a new column with the running total of passengers using the
`cumsum()` (cumulative sum) function.

```{r}

waypoints <- waypoints |>
  mutate(num_on_board = cumsum(name))

waypoints |>
  kable()

```

## QGIS

In QGIS, open the attribute table, start an editing session, and tap the 
field calculator button.

Check the box for "Create a new field" and enter a name for your new field.
Make sure the data type is numeric.

For the expression, type:

`with_variable (`
`    'time',`
`    time,`
`    sum( name, filter:=time<=@time)`
`)`

This will create a running total in the order of your time variable.

```{r, echo=FALSE, message=FALSE}
here("week3",
     "examples",
     "gifs",
     "qgis-cumsum.gif") |>
  knitr::include_graphics()
```

## ArcGIS

In ArcGIS, you will need to start by creating a new for the cumulative sum.
Make sure you set the field type to "Long" (for integers) or "Double" (for 
numbers with decimal places - which you don't need, but will still work).

```{r, echo=FALSE, message=FALSE}
here("week3",
     "examples",
     "gifs",
     "arcgis-new-field.gif") |>
  knitr::include_graphics()
```


Then, you'll need to create a function to calculate the running total. Open the
Calculate Field tool. Specify Input Table as the attribute table (in this example, it's the attributes table for my waypoints) and the name of the field
you just created that you want to calculate values for. 

In the box under "[name of your new field] =", type the name you're going to 
give your function (in this example, I use "runningTotal"), with the name 
of the variable you want to apply it to (this will probably be "Name" if you
are using your waypoint annotations) inside parentheses and exclamation points.

In the code block, paste the following lines: 

```{r, eval=FALSE}
total = 0
def runningTotal(value):
    global total
    total += value
    return total
```

    
This starts by setting a variable to the value of zero. Then it defines a
function (use the same name you used in the text box above it), and it 
sequentially adds the value from the other column to calculate a running 
total.

Once you've done that, click Run, and confirm that your column has been 
filled in with a running total.

```{r, echo=FALSE, message=FALSE}
here("week3",
     "examples",
     "gifs",
     "arcgis-make-sum-tool.gif") |>
  knitr::include_graphics()
```

# Assign point attributes to a line layer

Once you've calculated the passenger load that would follow each point, you
might need to transfer that variable to your line segments.

If you know that your waypoints and line segments are in the same order (i.e.
in order by time), this is called a positional join.

## R

In R, you can create a drop the geometry from the points layer and use 
`cbind()` to attach those columns to the line segments layer.

```{r}
 
point_data <- waypoints |>
  st_drop_geometry()

segments <- segments |>
  cbind(point_data)

segments |>
  kable()

```


## QGIS

QGIS doesn't do positional joins: you need to have a common column to match
values on. If both of your layers are in right order, you can just add a column
of sequential values and use that for the join. 

To add a column of sequential values to layer, select Toolbox from the 
Processing menu and select "Add autoincremental field". Select the input layer
you want to use and the number you want to start at (probably 1, but it doesn't
matter as long as you make the same choice for both layers). Then click
Run. This will create a new scratch layer with this additional field. Before
you move on (and certainly befor you close your project), you should make it
permanent.

```{r, echo=FALSE, message=FALSE}
here("week3",
     "examples",
     "gifs",
     "qgis-increment.gif") |>
  knitr::include_graphics()
```

Now you can join the data in the point layer to the data in the line layer. 
Right-click on the segments layer and select properties. From the properties
dialog, select Joins and click the green plus sign to add a join. Select
the layer you want to join data from and the names of the column you want to 
join by for both layers.

```{r, echo=FALSE, message=FALSE}
here("week3",
     "examples",
     "gifs",
     "qgis-join.gif") |>
  knitr::include_graphics()
```

## ArcGIS

ArcGIS doesn't do positional joins - you need a common column to do the join.
Your segments layer that you created by splitting the tracks at the points
already has a variable called ORIG_SEQ. In my example, this variable has the 
segments in sequence by time.

I need to create a variable for the waypoints that does the same. I'll create
a new field and use the field calculator to fill that field with sequential
numbers, using the following code in a code block to define a function
called `sequence()`:

```{r, eval=FALSE}
start = 0

def sequence():
    global start
    start += 1
    return start
```

```{r, echo=FALSE, message=FALSE}
here("week3",
     "examples",
     "gifs",
     "arcgis-add-order-field.gif") |>
  knitr::include_graphics()
```

Now I can join by points layer to the segments layer based on those two variables
defining the row sequence.

```{r, echo=FALSE, message=FALSE}
here("week3",
     "examples",
     "gifs",
     "arcgis-join.gif") |>
  knitr::include_graphics()
```
