---
title: "Skills: Spatial Data Calculations"
subtitle: "Week 3"
output: 
  rmdformats::readthedown:
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#knitr::knit_engines$set(python = reticulate::eng_python) 
```

This page introduces useful skills for performing calculations with
spatial data using R, QGIS, or ARcGIS.

The R skills on this page use the following R packages:

```{r, message=FALSE}
library(sf)
library(here)
library(tidyverse)
library(knitr)
library(crsuggest)
```

The examples here use the following data sets from the Open Data Portal for 
Bristol, United Kingdom:

* [Air Quality Management Areas](https://opendata.bristol.gov.uk/datasets/bcc::air-quality-management-areas-1/explore)
* [Public bike pumps](https://opendata.bristol.gov.uk/datasets/bcc::public-bike-pumps-1/explore)
* [Wards](https://opendata.bristol.gov.uk/datasets/bcc::wards/explore)
   * Some Wards (a subset of wards from the Wards layer)

```{r, echo=FALSE, message=FALSE, results='hide'}
aqma <- here("week3",
             "examples",
             "Air_Quality_Management_Areas") |>
  st_read()

pumps <- here("week3",
             "examples",
             "Public_bike_pumps.geojson") |>
  st_read()

wards <- here("week3",
             "examples",
             "Wards") |>
  st_read()

some_wards <- wards |>
  filter(NAME == "Ashley" |
           NAME == "Cotham" |
           NAME == "Eastville" |
           NAME == "Southville" |
           NAME == "Lockleaze")
```


# Checking a layer's coordinate reference system

If you'll be doing anything that involves measuring distances or 
areas, you should have your data in a projected coordinate system
(coordinates in feet or meters rather than degrees longitude and
latitude). So it's worth checking what coordinate reference system
your data are in.

## R

In R, you can check the coordinate reference system for a layer
using `st_crs()`.

Here, I can see that my layer of ward polygons is in the British National
Grid projection, which is a projected coordinate system.

```{r}
st_crs(wards)
```

My bike pump data is in WGS 84, which is a geographic coordinate 
system.

```{r}
st_crs(pumps)
```

## QGIS

In QGIS, there is a coordinate reference system associated with your
map/project that controls how the spatial layers are displayed, and there
is is a coordinate reference system associated with the each layer that
refers to how coordinates are stored in the file.

The CRS for the map/project is shown on a button on the bottom right
corner of your map window, and you can change the map's CRS by clicking
that button.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
library(here)

here("week3",
     "examples",
     "gifs",
     "qgis-proj-crs.gif") |>
  knitr::include_graphics()
```

When you add the first layer to an empty map window, the map's CRS will
change to match the projection of that layer. You can check the layer's 
CRS from the General Information tab of the layer properties window.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
library(here)

here("week3",
     "examples",
     "gifs",
     "qgis-check-crs.gif") |>
  knitr::include_graphics()
```

## ArcGIS

In ArcGIS, there is a coordinate reference system associated with your
map/project that controls how the spatial layers are displayed, and there
is is a coordinate reference system associated with the each layer that
refers to how coordinates are stored in the file. 

You can check the map's CRS from the Map Properties windo under
"Coordinate Systems."

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "arcgis-check-map-proj.gif") |>
  knitr::include_graphics()
```

You can check a layer's coordinate reference system from the layer
properties window, under "Source."

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "arcgis-check-layer-crs.gif") |>
  knitr::include_graphics()
```

# Changing a layer's coordinate reference system

Two reasons to change a layer's coordinate reference system are:

1. Your data are in a geographic coordinate system and you want to 
do distance and area calculations that require a projected coordinate
system.
2. You need have multiple layers, each in a different coordinate 
system, and you need to do an operation that requires them to
be in the same coordinate system.

Changing the layer's coordinate system edits the layer data by changing
the value of any coordinates. It would not necessarily change the 
way the data are displayed on a map.

## R

The `st_transform()` function can transform a dataset to a new
CRS.

I can use the `crs_suggest()` function to suggest a good projected
coordinate reference system for my bike pump data.

```{r}
pumps <- pumps |>
  st_transform(suggest_top_crs(pumps))
```

The package has suggested UTM Zone 30N. I can confirm that this is the new 
crs for the layer.

```{r}
st_crs(pumps)
```

I might prefer to set the CRS to a coordinate system that will be 
consistent with one of the other layers I'm using. I can use 
`st_crs()` for this as well.

```{r}
pumps <- pumps |>
  st_transform(st_crs(wards))
```

And now my pumps data is using the British National Grid, just like
my wards data.

```{r}
st_crs(pumps)
```

## QGIS

In QGIS, you should start by reprojecting your data to the 
CRS you want to tranform it to. Again, this does not change
your underlying data, just the way it's displayed. 

Click on the button on the bottom right corner of your map window that 
shows the current projection, and select a new projection. You can
select from a list of recently-used projections, or from a longer
list of pre-defined projections. You can use the search bar to search
for a projection by name.

In this example, I'm reprojecting from the British National Grid to 
UTM Zone 30N. If this is a reprojection you haven't done before, 
you might see a dialog asking you to select a transformation method.
In this case, you can just accept the default.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "qgis-reproject.gif") |>
  knitr::include_graphics()
```

Once you've reprojected the data, you can export a new version 
of the file that uses the new CRS. Right-click on the layer in your layers
pane, select "Export" and "Save Features As..." Select the file type and the CRS of the new layer. Save the transformed layer in a place where you'll
find it later (with the rest of your project files). When you add the 
transformed layer to your map, it will look exactly like the one you
started with (probably in a different color) because both layers
are being *displayed* in the map's CRS, but the coordinate data for each
version is saved in a different coordinate reference system.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "qgis-crs-transfomr.gif") |>
  knitr::include_graphics()
```

## ArcGIS

In ArcGIS, you should start by reprojecting your data to the 
CRS you want to tranform it to. Again, this does not change
your underlying data, just the way it's displayed. 

You can change the map's coordinate system from the Map Properties window,
under "Coordinate System."

In this example, I'm reprojecting from the British National Grid to 
UTM Zone 30N. 

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "arcgis-reproject.gif") |>
  knitr::include_graphics()
```

Use the Project geoprocessing tool to create a new version of the of the
layer in the coordinate reference system that matches the current map
projection.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "arcgis-transform.gif") |>
  knitr::include_graphics()
```

# Splitting polygons based on overlaps with another polygon layer

You might be concerned with the degree to which one (set of) polygon
overlaps with another (set of) polygon(s). You can start by splitting
a polygon into its overlapping and non-overlapping portions.

You've already worked with these methods a little if you've tried 
[clipping the water](https://c-voulgaris.github.io/VIS-2128/week2/skills.html#clipping-an-area-from-a-polygon) from a geographic area. 

## R

`st_intersection()` returns the part of a polygon that overlaps with
another layer, and `st_difference()` returns the part that doesn't
overlap.

Here is a plot showing the boundaries of the selected wards and the
boundary of the air quality management area.

```{r}
ggplot() +
  geom_sf(data = some_wards, fill = "lightblue", color = "blue") +
  geom_sf(data = aqma, fill = NA, color = "red") +
  theme_void()
```

I want to create a set of polygons representing the parts of these wards
that are within the air quality management area and another set that 
represents the parts of our selected wards that are outside the air quality management area.

```{r, warning=FALSE}
some_wards_in <- some_wards |>
  st_intersection(aqma)

some_wards_out <- some_wards |>
  st_difference(aqma)
```

And we can confirm that these polygons represent the areas inside 
and outside the AQMA boundary.

```{r}
ggplot() +
  geom_sf(data = some_wards_in, fill = "pink", color = "gray") +
  geom_sf(data = some_wards_out, fill = "powderblue", color = "gray") +
  geom_sf(data = aqma, fill = NA, color = "red") +
  theme_void()
```

## QGIS

Hover over the "Geoprocessing tools" item on the "Vector" menu. To create
a set of polygons representing the parts of a layer that are inside
the boundaries of another layer, use the "Intersection" tool.

The input layer is the layer you want to clip. The overlay layer is the one
defining the areas you want to keep.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "qgis-intersection.gif") |>
  knitr::include_graphics()
```

To create
a set of polygons representing the parts of a layer that are outside
the boundaries of another layer, use the "Difference" tool.

The input layer is the layer you want to clip. The overlay layer is the one
defining the areas you want to remove.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "qgis-difference.gif") |>
  knitr::include_graphics()
```

## ArcGIS

The Clip tool in ArcGIS can modify a layer to clip away everything
within another layer, to everything outside another layer, or to keep all
the polygons in a layer, but break them into the parts that are inside
and outside another layer.

Before you modify a layer, you may want to save a copy of it.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "arcgis-save-copy.gif") |>
  knitr::include_graphics()
```

Here is how you would use the Clip tool to keep everything outside another layer.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "arcgis-difference.gif") |>
  knitr::include_graphics()
```

Here is how you would keep everything inside another layer.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "arcgis-intersection.gif") |>
  knitr::include_graphics()
```

And here is how you split polygons into the components inside and outside
another layer, and keep all portions.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "arcgis-split.gif") |>
  knitr::include_graphics()
```

# Combining layers

You might have a set of features in two different layers that
you want to combine into a single layer.

## R

Continuing from our example where I creates a polygon layer
representing the parts of selected wards within the AQMA and another
polygon layer representing the areas outside the AQMA:

I can combine these two polygon layers into 
a single layer. Before I combine them, I'll create a variable to indicate
which is which.

```{r}
some_wards_in <- some_wards_in |>
  mutate(in_aqma = TRUE)

some_wards_out <- some_wards_out |>
  mutate(in_aqma = FALSE)
```

And then I'll use `rbind()` to combine the two tables. This requires
that the two data sets have the exact same set of columns (with the
same names, in the same order).

```{r}
some_wards_in_out <- rbind(some_wards_in, some_wards_out)
```

And now I can plot the resulting layer on a map.

```{r}
ggplot() +
  geom_sf(data=some_wards_in_out, color="white",
          aes(fill=in_aqma)) +
  scale_fill_brewer(type="qual",
                    name = "Relation to AQMA",
                    labels = c("Outside",
                               "Inside")) +
    geom_sf(data=aqma, fill=NA, color="black") +
  theme_void()
```

## QGIS

Continuing from our example where I creates a polygon layer
representing the parts of selected wards within the AQMA and another
polygon layer representing the areas outside the AQMA:

I can combine these two polygon layers into 
a single layer. Before I combine them, I'll create a variable to indicate
which is which.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "qgis-add-in-out.gif") |>
  knitr::include_graphics()
```

To merge two layers into a single layer, hover over "Data management tools"
from the "Vector" menu, and select "Merge Vector Layers...". Save the 
resulting file with the rest of your project files.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "qgis-merge layers.gif") |>
  knitr::include_graphics()
```

## ArcGIS

You may want to start by adding a variable to indicate which features come
from which layer.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "arcgis-in_out.gif") |>
  knitr::include_graphics()
```

And then you can merge the two layers. You'll need to indicate which 
variables you want the new, combined layer to include, and where to find
those variable values in each of the layers you're combining.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "arcgis-merge.gif") |>
  knitr::include_graphics()
```

# Finding the area of a polygon

Sometimes it's helpful to know the area of a polygon.

## R

`st_area` will calculate the area of a polygon, or each polygon 
within a layer of polygons. 

```{r}
some_wards_in_out <- some_wards_in_out |>
  mutate(area = st_area(some_wards_in_out)) |>
  select(NAME, area, in_aqma)

some_wards_in_out |>
  kable()
```

## QGIS

In QGIS, you can calculate the area from the field 
calculator.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "qgis-area-calc.gif") |>
  knitr::include_graphics()
```

## ArcGIS

In ArcGIS, you can right-click any heading on the attribute 
table and select "Calculate Geometry." 

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "arcgis-area-calc.gif") |>
  knitr::include_graphics()
```

# Summarizing values within categories

You might have features classified into multiple categories and you
want to calculate the total or average value of a variable within each 
category.

## R

The `group_by()` function will allow you to make calculations within 
categories

```{r}
some_wards_in_out <- some_wards_in_out |>
  group_by(in_aqma) |>
  mutate(total_area = sum(area))

some_wards_in_out |>
  kable()
```

## QGIS 

The `sum()` function in QGIS lets you specify a `group_by` parameter.
Here is how I would calculate the total area of each ward that 
each polygon in this layer belongs to.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "qgis-group-sums.gif") |>
  knitr::include_graphics()
```

## ArcGIS

In ArcGIS, you can use the Summarize button on the Table tab to create a
table that calculates summary statistics for a variable within categories
defined by another variable.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "arcgis-sum-by-group.gif") |>
  knitr::include_graphics()
```

# Creating a buffer

Some questions can be informed by the creation of a polygon representing
the area within a specified distance of a feature.

## R 

`st_buffer()` will create a buffer polygon around features in a layer.
The distance you specify should be in the same units as the coordinate
reference system.

Here is how you would create a set of circles defining a 20-meter radius
around each public bike pump.

```{r}
pump_buffer <- st_buffer(pumps, dist = 1000)
```

And here's how that looks on a map.

```{r}
ggplot() +
  geom_sf(data = pump_buffer, fill = "pink", color = "red") +
  geom_sf(data = pumps, color = "blue") +
  theme_void()
```

## QGIS

In QGIS you can create a buffer around a layer using the "Buffer"
geoprocessing tool. You can create a separate buffer around each 
feature by leaving the "Dissolve result" box unchecked.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "qgis-point-buff-sep.gif") |>
  knitr::include_graphics()
```

Or you can check the "Dissolve result" box to create a single feature representing
the area within a specified distance of *any* feature.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "qgis-point-buff-diss.gif") |>
  knitr::include_graphics()
```

## ArcGIS

The BUffer geoprocessing tool in ArcGIS will create a buffer around
features in a layer. If you set "Dissolve Type" to "No Dissolve", you'll
get one buffer for each individual feature.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "arcgis-buffer-sep.gif") |>
  knitr::include_graphics()
```

If you set "Dissolve Type" to "Dissolve all output features into a single feature", you'll
get one buffer for the entire area within a set distance of any feature in 
the input layer.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "arcgis-buffer-diss.gif") |>
  knitr::include_graphics()
```

# Combining features

You might have a set of features that you want to combine into
a single feature. This is called a *union*. A union will erase the
boundaries between overlapping or adjacent polygons, and it will create
a multipolygon to include non-adjacent polygons.

## R

Here's the set of buffers I generated around bike pumps in Bristol.

```{r}
ggplot() +
  geom_sf(data = pump_buffer, fill = "pink", color = "red") +
  theme_void()
```


I can use `st_union()` to combine them into a single multipolygon.

```{r}
one_pump_buffer = st_union(pump_buffer)
```

And here's the result.

```{r}
ggplot() +
  geom_sf(data = one_pump_buffer, fill = "pink", color = "red") +
  theme_void()
```

## QGIS

The "Union" geoprocessing tool in QGIS will combine polygons into a 
single, multipolygon, but it will preserve the boundaries
between them and it will keep overlapping polygons as overlapping parts
of the multipolygon.

If you want to dissolve the boundaries between polygons, you'll
need the "Dissolve" tool.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "qgis-dissolve.gif") |>
  knitr::include_graphics()
```

## ArcGIS

In ArcGIS, you can use the "Dissolve" geoprocessing tool to combine
all features in a layer, dissolving boundaries between adjacent and 
overlapping polygons.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "arcgis-dissolve.gif") |>
  knitr::include_graphics()
```

# Calculating the distance to the nearest feature

Calculating distances between things is the basis for most spatial data 
analysis.

## R

`st_distance()` will return the distance each of the features in 
another layer. Depending on what you're trying to do, it may be best
to find the distance to a layer that only includes a single feature. 
For a polygon (or multipolygon),
the distance will the distance to the closest part of the polygon 
(zero if the polygon contains the feature you're calculating the distance
from).

```{r}
pumps <- pumps |>
  mutate(distance = (st_distance(pumps, aqma))) |>
  select(distance)

head(pumps) |>
  kable()
```

```{r, warning=FALSE}
ggplot() +
  geom_sf(data = aqma) +
  geom_sf(data = pumps,
          aes(color = as.numeric(distance))) +
  scale_color_viridis_c(trans = "log",
                        name = "Distance from\nAQMA (meters)",
                        breaks = breaks <- 100*2^seq(0, 20, by=1),
                        labels = prettyNum(breaks, big.mark = " ")) +
  theme_void()
```

## QGIS

The NNjoin plugin for QGIS makes it easy to calculate the distance from 
features in one layer to the nearest features in another layer. Before 
you can use it, you need to install it.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "qgis-install-nnjoin.gif") |>
  knitr::include_graphics()
```

Once you've installed the plugin, a yellow NNjoin button will appear on your
screen. Click this button and a dialog will appear where you can specify
an input layer, a join layer, and an output layer. The output layer
will be a copy of the input layer that will include column from the join
layer, including a column indicating the distance to the nearest feature.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "qgis-calc-dist.gif") |>
  knitr::include_graphics()
```

## ArcGIS

The Near tool in the Analysis Toolbox will calculate the distance to the
nearest feature in another layer. It will add a field to the attribute 
table called "NEAR_DIST" (by default - you can change this) that will indicate
the distance to the nearest feature in the layer you've specified.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week3",
     "examples",
     "gifs",
     "arcgis-dist-calc.gif") |>
  knitr::include_graphics()
```