---
title: "Skills: Digitizing historic map data"
subtitle: "Week 6"
output: 
  rmdformats::readthedown:
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#knitr::knit_engines$set(python = reticulate::eng_python) 
```

This page introduces useful skills for extracting spatial data from 
historic maps

# Creating a georeferenced raster layer from an image

If you have an image file from a scanned map (for example, a historic
map, or one you or a collaborator drew/annotated by hand), you can
import it into GIS software and georeference it, which associates each
pixel in your image with a geospatial location.

In QGIS, you may want to start by loading a basemap that you can
use to identify locations on the map you're georeferencing. You can
add a basemap based on OpenStreetMap data by going to "Add Layer" from
the "Layer" menu and selecting "Add XYZ layer." In the dialog that 
comes up, you'll be able to select "OpenStreetMap" from the 
dropdown menu. Then click the "Add" button

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
library(here)

here("week6",
     "Examples",
     "gifs",
     "qgis-add-osm-base.gif") |>
  knitr::include_graphics()
```

Now you can georeference your image. Select "Georeferencer..."
from the layer menu. 

In the georeferencer dialog, click the "Open Raster" button to load
your image file. You'll need to select at least three points on the
image to georeference it. Ideally, there will be spaced far apart
from one another (e.g. near the sides of the image) and will form 
a triangle (rather than being along a straight line).

Click the first point on your image. A box will pop up asking you
for the coordinates of the point you've selected. Click the 
"From Map Canvas" button and find the corresponding location on
in your map window. Click on the location and click okay.

Repeat this process for at least two more points. When you are 
satisfied with your control points, click the Settings button. 
Enter a filename for the georeferenced image and make sure the 
"Load in project when done" button is checked. Click Okay.

Click the "Start Georeferencing" button. When the process is finished, 
you can close the georeferencer dialog box. You'll be prompted to save 
control points, which isn't necessary.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week6",
     "Examples",
     "gifs",
     "qgis-georef.gif") |>
  knitr::include_graphics()
```

# Manually creating a layer of data

Once you've loaded a georeferenced layer, you might want to features
to create a vector layer.

In QGIS, you'll want to start by creating an empty shapefile. 
Go to Layer > Create Layer > New Shapefile Layer...

Select a filename and a geometry type, then click okay. The new, 
empty layer will be added to your map.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week6",
     "Examples",
     "gifs",
     "qgis-make-shp.gif") |>
  knitr::include_graphics()
```

Now you'll want to add features to your new layer. Right-click the 
layer you want to add to (probably the empty layer you just
created) and select "Toggle Editing". Then click the "Add points"
button and click on map where you want your new points to be. When
you are finished, save your edits and turn off the editing mode.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week6",
     "Examples",
     "gifs",
     "qgis-add-points.gif") |>
  knitr::include_graphics()
```

