---
title: "Skills: Collecting and importing field data"
subtitle: "Week 5"
output: 
  rmdformats::readthedown:
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#knitr::knit_engines$set(python = reticulate::eng_python) 
```

```{r, echo=FALSE, message=FALSE}
library(here)
```


This page demonstrates methods for collecting GPS data in the field and importing
it to a GIS software of your choice.

# Collecting GPS traces

There are many different smart phone applications for both Andorid
and iOS devices that will allow you to collect GPS traces and export
them as GPX files. These include:

* Strava ([iOS](https://apps.apple.com/us/app/strava-run-bike-hike/id426826309) and [Android](https://play.google.com/store/apps/details?id=com.strava&hl=en_US&gl=US)): A social app for sharing data on workouts, including
running, cycling, hiking, canoeing, skiing, etc. You can record and share the
route you took on a "workout" and export your route as a GPX file.
* OsmAnd ([iOS](https://apps.apple.com/us/app/osmand-maps-travel-navigate/id934850257) and [Android](https://play.google.com/store/apps/details?id=net.osmand&hl=en_US&gl=US)): A navigation app build on OpenStreetMap that also
will record GPX traces if you enable the "Trip recording" plugin.
* GPX Tracker ([iOS](https://apps.apple.com/us/app/open-gpx-tracker/id984503772)): A specialized app that just records GPX traces.

The demonstration below uses the OsmAnd app for iOS.

When you first install and open the app, you'll see a screen like this:

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "IMG_0489.PNG") |>
  knitr::include_graphics()
```


```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "IMG_0490.PNG") |>
  knitr::include_graphics()
```


The first screen will ask you to download a map that corresponds with your 
current location. This is to allow the app to work offline (without an 
internet connection). You can skip downloading the map and still record 
GPS locations, but the app won't show you a basemap in the background.

Once you've downloaded a basemap, the download button becomes a "Go to the map"
button. Tap that button.

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "IMG_0491.PNG") |>
  knitr::include_graphics()
```

You'll see a map of your current surroundings. On the botton right corner of the map, there
will be a little crosshair icon. If you tap that, you'll get a message asking you to give
the app permission to use your location. 

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "initial-map.jpeg") |>
  knitr::include_graphics()
```

Tracking your location is the whole point of this exercise. Choose "Allow While Using App" if you
intend to keep the app open the whole time you're collecting data. Otherwise, you want to set it
to track your location all the time.

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "IMG_0503.PNG") |>
  knitr::include_graphics()
```

Tap the little hamburger button on the bottom left corner of the map to open up the side menu. 

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "IMG_0492.PNG") |>
  knitr::include_graphics()
```

Tap "My places". Select the "Tracks" icon on the bottom right and tap the banner that says,
"Please activate Track recording addon to record and save your tracks."

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "IMG_0494.PNG") |>
  knitr::include_graphics()
```


You'll see a list of plugins with white checkboxes next to them. 

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "unchecked-trip-rec.jpeg") |>
  knitr::include_graphics()
```

Tap the box by "Trip Recording." It will show you a message saying you've added the plugin. Click OK.

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "IMG_0495.PNG") |>
  knitr::include_graphics()
```

Now the checkbox will be highlighted in orange. Tap on "< My Tracks". 

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "IMG_0496.PNG") |>
  knitr::include_graphics()
```

You'll see a banner that says "Currently recording track" with a gray save icon and
a red dot. Tap the red dot. 

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "IMG_0498.PNG") |>
  knitr::include_graphics()
```

A dialog will pop up where, if you want, you can 
adjust the recording interval. By default, it will record your location every five 
seconds. Click OK.

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "IMG_0497.PNG") |>
  knitr::include_graphics()
```

Now you should see a black square instead of a red circle. The app is currently recording your
location. Start walking along the path you want to record. When you have finished your path, 
tap the black square to stop recording.

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "IMG_0505.PNG") |>
  knitr::include_graphics()
```

Tap the "Rec" folder, and you'll see your track listed with a file name
indicating the current date and time.

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "IMG_0499.PNG") |>
  knitr::include_graphics()
```

Tap the name of your track and it will open in a map.

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "IMG_0500.PNG") |>
  knitr::include_graphics()
```

Tap the share icon at the bottom of the screen and email the GPX file to yourself.

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "IMG_0501.PNG") |>
  knitr::include_graphics()
```

# Taking geotagged photos

For the most part, your phone is collecting the latitude/longitude
coordinates of each photo you take by default, unless you've
explicitly turned off location access for your phone's camera. 

# Importing GPS traces

Coming soon...

# Extracting locations from geotagged photos

Coming soon...

# Importing csv locatations

Coming soon...


