---
title: "Skills: Collecting and importing field data"
subtitle: "Week 5"
output: 
  rmdformats::readthedown:
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#knitr::knit_engines$set(python = reticulate::eng_python) 
```


This page demonstrates methods for collecting GPS data in the field and importing
it to a GIS software of your choice.

The R methods on this page use the following packages:


```{r, message=FALSE}
library(here)
library(sf)
library(tidyverse)
library(ggspatial)
library(exifr)
```

# Collecting GPS traces

There are many different smart phone applications for both Andorid
and iOS devices that will allow you to collect GPS traces and export
them as GPX files. These include:

* Strava ([iOS](https://apps.apple.com/us/app/strava-run-bike-hike/id426826309) and [Android](https://play.google.com/store/apps/details?id=com.strava&hl=en_US&gl=US)): A social app for sharing data on workouts, including
running, cycling, hiking, canoeing, skiing, etc. You can record and share the
route you took on a "workout" and export your route as a GPX file.
* OsmAnd ([iOS](https://apps.apple.com/us/app/osmand-maps-travel-navigate/id934850257) and [Android](https://play.google.com/store/apps/details?id=net.osmand&hl=en_US&gl=US)): A navigation app build on OpenStreetMap that also
will record GPX traces if you enable the "Trip recording" plugin.
* GPX Tracker ([iOS](https://apps.apple.com/us/app/open-gpx-tracker/id984503772)): A specialized app that just records GPX traces.

The demonstration below uses the OsmAnd app for iOS (with some examples from an Android
device when things look different).

When you first install and open the app, you'll see a screen like this:

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "IMG_0489.PNG") |>
  knitr::include_graphics()
```

The first screen will ask you to download a map that corresponds with your 
current location. This is to allow the app to work offline (without an 
internet connection). You can skip downloading the map and still record 
GPS locations, but the app won't show you a basemap in the background.

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "IMG_0490.PNG") |>
  knitr::include_graphics()
```

If your phone asks apps not to track your location by default, the app might guess the wrong
map to download for your current location. In this case, you can tap "Skip Download" on the
bottom left corner of the screen.

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "image000000-ct.png") |>
  knitr::include_graphics()
```

Then you can search for the map you want to download.

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "image000000 (1).png") |>
  knitr::include_graphics()
```

Once you've downloaded a basemap, the download button becomes a "Go to the map"
button. Tap that button.

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "IMG_0491.PNG") |>
  knitr::include_graphics()
```

You'll see a map of your current surroundings. On the botton right corner of the map, there
will be a little crosshair icon. If you tap that, you'll get a message asking you to give
the app permission to use your location. 

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "initial-map.jpeg") |>
  knitr::include_graphics()
```

Tracking your location is the whole point of this exercise. Choose "Allow While Using App" if you
intend to keep the app open the whole time you're collecting data. Otherwise, you want to set it
to track your location all the time.

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "IMG_0503.PNG") |>
  knitr::include_graphics()
```

Tap the little hamburger button on the bottom left corner of the map to open up the side menu. 

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "IMG_0492.PNG") |>
  knitr::include_graphics()
```

Tap "My places". Select the "Tracks" icon on the bottom right and tap the banner that says,
"Please activate Track recording addon to record and save your tracks."

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "IMG_0494.PNG") |>
  knitr::include_graphics()
```


On an Android phone, you may need to add this plugin from the Plugin manager
(find it towards the bottom of the side menu, right before settings).

On an iPhone, you'll see a list of plugins with white checkboxes next to them. 

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "unchecked-trip-rec.jpeg") |>
  knitr::include_graphics()
```

On an Android, it will look like this:

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "image000000 (2).png") |>
  knitr::include_graphics()
```

Tap the box or icon by "Trip Recording." It will show you a message saying you've added the plugin. Click OK.

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "IMG_0495.PNG") |>
  knitr::include_graphics()
```

Now the checkbox or icon will be highlighted in orange. Tap on "< My Tracks". 

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "IMG_0496.PNG") |>
  knitr::include_graphics()
```

You'll see a banner that says "Currently recording track" with a gray save icon and
a red dot. Tap the red dot. 

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "IMG_0498.PNG") |>
  knitr::include_graphics()
```


On an Android device, the "Start Recording" button is somewhat more 
clearly labeled.


```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "image000000[3].PNG") |>
  knitr::include_graphics()
```


A dialog will pop up where, if you want, you can 
adjust the recording interval. By default, it will record your location every five 
seconds. Setting it to a shorter interval (e.g. every one second) will improve the
precision of your GPS trace.

Click OK.

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "IMG_0497.PNG") |>
  knitr::include_graphics()
```

Now you should see a black square instead of a red circle. The app is currently recording your
location. Start walking along the path you want to record. When you have finished your path, 
tap the black square to stop recording. Then tap the save icon.

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "IMG_0505.PNG") |>
  knitr::include_graphics()
```

On an Android device the start, stop, and save icons are more clearly labeled.

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "image000000.PNG") |>
  knitr::include_graphics()
```


Tap the "Rec" folder, and you'll see your track listed with a file name
indicating the current date and time.

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "IMG_0499.PNG") |>
  knitr::include_graphics()
```

Tap the name of your track and it will open in a map.

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "IMG_0500.PNG") |>
  knitr::include_graphics()
```

Tap the share icon at the bottom of the screen and email the GPX file to yourself.

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "IMG_0501.PNG") |>
  knitr::include_graphics()
```

To share the file from an Android device, you'll need to click the "Options"
button on the bottom left corner of your screen.

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "image000000[1].png") |>
  knitr::include_graphics()
```

Then click the Share item on the side menu.

```{r, echo=FALSE, message=FALSE, out.width = '30%'}
here("week5",
     "screenshots",
     "image000000[2].png") |>
  knitr::include_graphics()
```

# Taking geotagged photos

For the most part, your phone is collecting the latitude/longitude
coordinates of each photo you take by default, unless you've
explicitly turned off location access for your phone's camera. 

# Importing GPS traces

Once you've exported a GPX file from a GPS device (such as a smart phone),
you may want to bring it into your GIS software.

GPX files include data in five layers:

* route_points
* routes
* track_points
* tracks
* waypoints

If you've just collected a GPS trace using an app like Strava, GPX Tracker, or OsmAnd,
the app was recording your position and the current time at regular intervals (e.g.
every second, or every five seconds). Those times and locations are stored in the
track_points layer. They are connected into a line (based on the order in which they
were recorded) that is stored in the tracks layer.

## R

You can use the `st_read()` function to load data from a GPX file into
R. You will need to specify which layer you want.

```{r, message=FALSE, results='hide'}
path_points <- here("week5",
     "examples",
     "04-Aug-2023-1332.gpx") |>
  st_read(layer = "track_points")

path_line <- here("week5",
     "examples",
     "04-Aug-2023-1332.gpx") |>
  st_read(layer = "tracks")
```

And here are the points and the associated line displayed on a map. Note that 
I'm using `annotation_map_tile` to display a base map under my layers.

It would likely be sufficient to display the line without the points, but viewing 
the points could be useful for your own information. For example, in this example, 
the line makes it look like I cut through a building, when what actually happened
is that my device stopped recording my location for a few seconds while I walked
around the corner.

```{r}
ggplot() +
  annotation_map_tile(type = "cartolight",
                      zoomin = 0,
                      quiet=TRUE,
                      progress = "none") +
  geom_sf(data = path_line) +
  geom_sf(data = path_points,
          alpha = 0.2) +
  theme_void()
```

## QGIS

You can import a GPX file to QGIS by dragging and dropping it onto your map.
You'll be able to specify which layers you want. The most relevant layers are
likely to be "track_points" and "tracks".

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week5",
     "examples",
     "gifs",
     "qgis-import.gif") |>
  knitr::include_graphics()
```

## ArcGIS

You can use the GPX to Features tool (in the Geoprocessing toolbox)
in import a GPX file to ArcGIS Pro. If you want to import both points
and lines, you'll need to import them separately.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week5",
     "examples",
     "gifs",
     "arcgis-import.gif") |>
  knitr::include_graphics()
```

# Extracting locations from geotagged photos

Most digital cameras record metadata about each picture file as EXIF (Exchangeable
Image File) data, which often includes the latitude/longitude coordinates for 
where the picture was taken.

## Web browser

There a websites that will display EXIF data for photos that you upload. One
example is https://onlineexifviewer.com/

Once you upload a photo, it will display all the associated metadata, including
the location where the photo was taken. You can copy these coordinates to your
clipboard and paste them into a text file or spreadsheet to create a csv file.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week5",
     "examples",
     "gifs",
     "exif-web.gif") |>
  knitr::include_graphics()
```

## R

The `exifr` package will import EXIF data from a set of photos into
a data frame in R.

`exifr` is actually a wrapper around a Perl program (Perl is a programming language), 
so you'll need to install Perl in order for it to run (before you load the package). 
On Windows, you can install perl from
https://strawberryperl.com/. On a Mac, you may already have perl installed.

Once you've installed Perl and the `exif` package, you can load the `exif` package. 
Then, run the function `configure_exiftool()` in the command line. You only need 
to do this once. Then you can run the `read_exif()` function to extract the coordinates
for all the photos in a directory by doing something like this:

```{r}
photo_data <- here("week5",
                   "examples",
                   "photos") |>
  list.files(full.names = TRUE) |>
  read_exif(tags = c("GPSLatitude", "GPSLongitude")) 
```

The result will be a data frame with three columns: 

* `SourceFile`: The full path to the file on your computer
* `GPSLatitude`: The latitude coordinate for where the photo was taken
* `GPSLongitude`: The longitude coordinate for where the photo was taken

You can save this as a csv file that you read into another GIS software.

```{r}
write_csv(photo_data,
          file = here("week5",
                      "examples",
                      "photo-locs.csv"))
```

Or you can convert it into a set of points to work with in R.

```{r}
photo_points <- photo_data |>
  st_as_sf(coords = c("GPSLongitude", "GPSLatitude"),
           crs = "WGS84")

ggplot() +
  annotation_map_tile(type = "cartolight",
                      zoomin = 0,
                      quiet=TRUE,
                      progress = "none") +
  geom_sf(data = photo_points,
          color = "purple") +
  theme_void()
```

# Importing csv locations

If you have a csv file with latitude/longitude coordinates for a set of points,
you can easily import that data to any GIS software.

## R

If you read a csv file into R as a data frame, you can use `st_as_sf` to 
convert that to a spatial data layer. You'll need to specify the column
names for the x and y coordinates and the coordinate reference system.

```{r}
csv_points <- here("week5",
     "examples",
     "photo-locs.csv") |>
  read_csv(show_col_types = FALSE) |>
  st_as_sf(coords = c("GPSLongitude", "GPSLatitude"),
           crs = "WGS84")

ggplot() +
  annotation_map_tile(type = "cartolight",
                      zoomin = 0,
                      quiet=TRUE,
                      progress = "none") +
  geom_sf(data = csv_points,
          color = "orange") +
  theme_void()
```

## QGIS

In QGIS, go to "Add Layer" from the "Layer" menu and choose 
"Add delimited text layer." Navigate to the csv file you want to import,
specify the columns containing the x and y values, and indicate the 
coordinate reference system those coordinates refer to.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week5",
     "examples",
     "gifs",
     "qgis-csv-points.gif") |>
  knitr::include_graphics()
```

## ArcGIS

To add points from a csv file, click the "Add Data" button and then "XY Point Data".

This will open the XY Table to Point tool. Navigate to the csv file, speficy the columns
with x and y coordinates and coordinate system.

```{r, echo=FALSE, message=FALSE, out.width = '100%'}
here("week5",
     "examples",
     "gifs",
     "arcgis-csv-points.gif") |>
  knitr::include_graphics()
```

